### 2024-02-13 paper/paper.qmd
Update paper.qmd

---
title: "Red States = Red Flags?"
subtitle: "An analysis of birth rate trends from 2001 to 2019 and an exploration of political affiliation of different states and Birth rates in the United States,"
author: 
  - Rayan Awad Alim
  - Maria Mangru
  - Isha Junega
thanks: "Code and data are available at: https://github.com/RayanAlim/Birth_Rate_Analysis"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

Introduction

broader context to motivate The birth rate is one of the most important determinants of a country’s population growth, potentially impacting policy decisions about the health system, education, and economy (1). https://www.waldenu.edu/programs/health/resource/why-birth-rates-are-significant-in-health-studies-and-health-science

some detail about what the paper is about This paper dives into the potential factors that may cause a birth rate to decrease. a clear gap that needs to be filled Not sure what was done We replicated the paper by Kearney, Levine, andPardue (2022) and focused on wages, access to healthcare (with religious importance) and childcare expenditure data to draw conclusions about whether these factors affect the decline or birth rates what was found

why is it important the structure of the paper

Data

Article:

Model

Model InformationDependent variable: birth rate per 1000 populationIndependent variables:State min wage Abortion delay Child enforcement expenditure Political affiliation Education

Model set-up

Model defined as:

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the rstanarm package of @rstanarm. We use the default priors from rstanarm.

Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

Results

Our results are summarized in @tbl-modelresults.

{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))

1- Medicaid coverage has increased which allows more low-income people to get an abortion.

2- Maximum well-fare has gone down so it more difficult to raise a kid

3- Parents seem to have more of an interest in having an abortion which aligns with the decrease in number of births

{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)

Analysis Results

### Analysis Results

# State-Specific Birth Averages (2001-2019):

The average number of births varies considerably from state to state, with California (CA) having the highest average number of births and Vermont (VT) having the lowest.

Larger states, both in terms of geography and population like Texas (TX) and New York (NY), show significantly higher average numbers of births, which aligns with the expected outcomes based on population.

Smaller states such as Vermont (VT) and North Dakota (ND) have lower average numbers of births, which may reflect not just smaller populations but also demographic trends such as aging populations.

{r}
#INTREST DISTRO OVER TIME GRAPHS

# State Minimum Wage Changes (2001-2019):

The District of Columbia (DC) exhibits the largest increase, with its minimum wage rising by $7.85, from $6.15 in 2001 to $14.00 in 2019. Arizona (AZ) and Colorado (CO) also show substantial increases of $6.85 each.

Georgia (GA) and Wyoming (WY) stand out as the only states with no change in minimum wage, maintaining a rate of $5.15 across the years. It is worth noting that this rate is equal to the federal minimum wage between 1997 - 2007, suggesting that these states have not adjusted their minimum wage above the federal floor set at that time.

{r}
# MINIMUM WAGE GRAPH

# State-Specific Births by Abortion Delay Policy from 2001 to 2019:

The trend lines across the graphs indicate a general decline in birth rates. The decline is more pronounced in states that have implemented delay policies, suggesting a possible correlation between policy enactment and a faster reduction in birth rates.

States with Delay

Birth rates vary significantly between states. For example, Arkansas (AR) shows a sharp decrease in birth rates compared to other states. In contrast, Pennsylvania (PA) has a gentler slope.

The range of birth rates in 2001 spans from approximately 17 per 1000 (excluding Utah (UT)) people in certain states to below 15 per 1000 in others, with a convergence of rates around 12 to 14 per 1000.

The states in this graph are ones which had a policy change from no delay to delay which allows for the analysis of the before-and-after effects of policy implementation.

For example, Arizona's (AZ) implementation year of 2009 shows birth rate descends sharply afterward. In contrast, North Carolina (NC) in 2011 didn't show such an immediate shift.

The states without delay policies showcase a natural decline in birth rates over time, without the potential influence of policy changes. This trend is consistent across the states, with some variations in the steepness of the decline.

For example, California (CA) maintains a relatively high birth rate over the years, starting near 15 per 1000 in 2001 and ending around 12 per 1000 in 2019. This could be influenced by its socio-economic profile and immigration patterns.

All states show a general decline in relative birth rates over the 19-year period, which is in line with the national trend of declining birth rates.

State-Specific Trends:

California (CA): The trend in California shows a relatively steady decline throughout the period, without any sharp changes.

Florida (FL): Florida's trend is similar to California's, with a gradual decline.

North Carolina (NC): North Carolina, with a policy change in 201, shows a relatively stable trend before the policy change and a sharper decline after.

Texas (TX): Texas exhibited a noticeable decline after the policy implementation year of 2003.

When contrasting the birth rates of Texas and California, two states with large populations, we see that Texas's birth rates declined more sharply after its policy change in 2003 compared to California. This might suggest an association between the policy implementation and the accelerated decline in birth rates in Texas.

<!-- -->

Comparing Georgia and North Carolina, both of which implemented delay policies. Georgia's birth rates were rising before the policy change and then started declining after, while North Carolina's birth rates were relatively stable before 2011 and then declined.

Discussion

## First discussion point: Abortion

{r}
#relative average # of births in states delay

#relative average # of births in states w/o delay

#relative birth rates in selected states

Second discussion: Political affiliation

{r}
# POLITICAL AFFILIATION MAP

Third discussion point: Minimum Wage

Fourth discussion point: Education

{r}
#Education graph

Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

Appendix

Additional data details

Model details

Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()

Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")

\newpage

References

### 2024-02-13 paper/Untitled.Rmd
update qmd and pdf

---
title: "regression_model"
output: html_document
date: "2024-02-13"
---

library(readxl)
library(readr)
library(haven)
library(tidyverse)

# Reading the datasets
abortion_policy <- read_dta("144981-V1/annual_policy/policyvars01_19.dta")
numbirths <- read_csv("144981-V1/annual_policy/numbirths_2001_2019.csv")
population_data <- read_excel("144981-V1/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")

# Filtering columns and filtering by year
abortion_policy_filtered <- abortion_policy %>%
  select(year, stname, delay, aca_mcaid_exp, pcthsdrop, pcthsgrad, pctsomecol) %>%
  filter(year >= 2001 & year <= 2019)

population_data_filtered <- population_data %>%
  select(state_name, year, 
         Governor_is_Democrat = `Governor is Democrat (1=Yes)`, 
         State_Minimum_Wage = `State Minimum Wage`) %>%
  filter(year >= 2001 & year <= 2019) %>%
  mutate(Governor_is_Democrat = na_if(Governor_is_Democrat, '.'))

numbirths_filtered <- numbirths %>%
  filter(year >= 2001 & year <= 2019)

# Merging data sets and removing rows with NA values
merged_data <- numbirths_filtered %>%
  left_join(abortion_policy_filtered, by = c("year", "stname" = "stname")) %>%
  left_join(population_data_filtered, by = c("stname" = "state_name", "year")) %>%
  drop_na(Governor_is_Democrat, delay, aca_mcaid_exp, pcthsdrop, pcthsgrad, pctsomecol, numbirth1544, State_Minimum_Wage)

# Running the regression model on the cleaned and filtered dataset
model <- lm(numbirth1544 ~ delay + aca_mcaid_exp + pcthsdrop + pcthsgrad + pctsomecol + Governor_is_Democrat + State_Minimum_Wage, data = merged_data)

# Summary of the model
summary(model)

###### Checking assumptions ######
install.packages("car")
install.packages("lmtest")
install.packages("sandwich")

library(tidyverse)
library(car)
library(lmtest)
library(sandwich)

# Check for linearity and homoscedasticity
plot(model, which = 1:2)

# Check for normality of residuals with a Q-Q plot
qqPlot(model, main = "Q-Q Plot")

# Test for independence of residuals (Durbin-Watson test)
dwtest(model)

# Check for multicollinearity with Variance Inflation Factors (VIF)
vif(model) # A VIF value > 5 or 10 indicates high multicollinearity

# Check for homoscedasticity with Breusch-Pagan test
bptest(model)

# Check for non-random residuals
# Residuals vs. Fitted values plot
plot(model, which = 1)
abline(h = 0, col = "red")

### 2024-02-13 scripts/demographic_data.R
Update demographic_data.R

@@ -32,10 +32,7 @@ ggplot(average_births_by_state, aes(y = reorder(stname, AverageBirths), x = Aver
   labs(title = "Average Number of Births by State (2001-2019)",
        y = "State",
        x = "Average Number of Births") +
-  theme(axis.text.y = element_text(angle = 0)) # Ensure the y-axis labels are horizontal
-
-# Note: Customize the aesthetics as necessary to fit your dataset's specifics
-
+  theme(axis.text.y = element_text(angle = 0)) 
 
 
 

### 2024-02-13 paper/paper.qmd
update qmd and pdf
changes to pdf and scripts
Added graphs to paper output and changed the path for various scripts

@@
 
-{r}
+```{r echo=FALSE, message=FALSE}
 #| include: false
@@
 
+```
+
 Introduction
 
-broader context to motivate The birth rate is one of the most important determinants of a country’s population growth, potentially impacting policy decisions about the health system, education, and economy (1).
+broader context to motivate The birth rate is one of the most important
+determinants of a country’s population growth, potentially impacting
+policy decisions about the health system, education, and economy (1).
 https://www.waldenu.edu/programs/health/resource/why-birth-rates-are-significant-in-health-studies-and-health-science
 
-some detail about what the paper is about This paper dives into the potential factors that may cause a birth rate to decrease.
-a clear gap that needs to be filled Not sure what was done We replicated the paper by Kearney, Levine, andPardue (2022) and focused on wages, access to healthcare (with religious importance) and childcare expenditure data to draw conclusions about whether these factors affect the decline or birth rates what was found
+some detail about what the paper is about This paper dives into the
+potential factors that may cause a birth rate to decrease. a clear gap
+that needs to be filled Not sure what was done We replicated the paper
+by Kearney, Levine, andPardue (2022) and focused on wages, access to
+healthcare (with religious importance) and childcare expenditure data to
+draw conclusions about whether these factors affect the decline or birth
+rates what was found
 
@@
 
-Model InformationDependent variable: birth rate per 1000 populationIndependent variables:State min wage Abortion delay Child enforcement expenditure Political affiliation Education
+Model Information Dependent variable: birth rate per 1000 population
+Independent variables:State min wage, Abortion delay, Child enforcement
+expenditure, Political affiliation
+
+Education
 
@@
 
+```{=tex}
 \begin{align} 
@@
 \end{align}
-
-We run the model in R [@citeR] using the rstanarm package of @rstanarm. We use the default priors from rstanarm.
+```
+We run the model in R [@citeR] using the rstanarm package of @rstanarm.
+We use the default priors from rstanarm.
 
@@
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+We expect a positive relationship between the size of the wings and time
+spent aloft. In particular...
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+We can use maths by including latex between dollar signs, for instance
+$\theta$.
 
@@
 
-{r}
+```{r echo=FALSE, message=FALSE}
 #| echo: false
@@
 
-1- Medicaid coverage has increased which allows more low-income people to get an abortion.
+```
+
+1- Medicaid coverage has increased which allows more low-income people
+to get an abortion.
 
@@
 
-3- Parents seem to have more of an interest in having an abortion which aligns with the decrease in number of births
+3- Parents seem to have more of an interest in having an abortion which
+aligns with the decrease in number of births
 
-{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
+{r} #\| echo: false #\| eval: true #\| label: tbl-modelresults #\|
+tbl-cap: "Explanatory models of flight time based on wing width and wing
+length" #\| warning: false
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
+modelsummary::modelsummary( list( "First model" = first_model ),
+statistic = "mad", fmt = 2 )
 
@@
 
-The average number of births varies considerably from state to state, with California (CA) having the highest average number of births and Vermont (VT) having the lowest.
+The average number of births varies considerably from state to state,
+with California (CA) having the highest average number of births and
+Vermont (VT) having the lowest.
 
-Larger states, both in terms of geography and population like Texas (TX) and New York (NY), show significantly higher average numbers of births, which aligns with the expected outcomes based on population.
+Larger states, both in terms of geography and population like Texas (TX)
+and New York (NY), show significantly higher average numbers of births,
+which aligns with the expected outcomes based on population.
 
-Smaller states such as Vermont (VT) and North Dakota (ND) have lower average numbers of births, which may reflect not just smaller populations but also demographic trends such as aging populations.
+Smaller states such as Vermont (VT) and North Dakota (ND) have lower
+average numbers of births, which may reflect not just smaller
+populations but also demographic trends such as aging populations.
 
-{r}
-#INTREST DISTRO OVER TIME GRAPHS
+```{r echo=FALSE, message=FALSE}
+library(haven)
+library(tidyverse)
+library(tidyr)
+
+# Importing the CSV files
+numbirths_2001_2019 <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv")
+births_educ_race_age6 <- read_csv("../data/analysis_data/decomp/births_educ_race_age6.csv")
+numbirths_educ_2044 <- read_csv("../data/analysis_data/educ/numbirths_educ_2044.csv")
+state_births_04_19 <- read_csv("../data/analysis_data/map/state_births_04_19.csv")
+
+# Display the first few rows of each data-frame to verify the imports
+view(numbirths_2001_2019)
+view(births_educ_race_age6)
+view(numbirths_educ_2044)
+view(state_births_04_19)
+
+### Number of Births ### 
+colnames(numbirths_2001_2019)
+
+average_births_by_state <- numbirths_2001_2019 %>%
+  filter(year >= 2001, year <= 2019) %>%
+  group_by(stname) %>%
+  summarize(AverageBirths = mean(numbirth1544, na.rm = TRUE)) %>%
+  ungroup()
+view(average_births_by_state)
+
+ggplot(average_births_by_state, aes(y = reorder(stname, AverageBirths), x = AverageBirths)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  theme_minimal() +
+  labs(title = "Average Number of Births by State (2001-2019)",
+       y = "State",
+       x = "Average Number of Births") +
+  theme(axis.text.y = element_text(angle = 0)) 
+```
 
@@
 
-The District of Columbia (DC) exhibits the largest increase, with its minimum wage rising by $7.85, from $6.15 in 2001 to $14.00 in 2019.
-Arizona (AZ) and Colorado (CO) also show substantial increases of $6.85 each.
+The District of Columbia (DC) exhibits the largest increase, with its
+minimum wage rising by \$7.85, from \$6.15 in 2001 to \$14.00 in 2019.
+Arizona (AZ) and Colorado (CO) also show substantial increases of \$6.85
+each.
 
-Georgia (GA) and Wyoming (WY) stand out as the only states with no change in minimum wage, maintaining a rate of $5.15 across the years.
-It is worth noting that this rate is equal to the federal minimum wage between 1997 - 2007, suggesting that these states have not adjusted their minimum wage above the federal floor set at that time.
+Georgia (GA) and Wyoming (WY) stand out as the only states with no
+change in minimum wage, maintaining a rate of \$5.15 across the years.
+It is worth noting that this rate is equal to the federal minimum wage
+between 1997 - 2007, suggesting that these states have not adjusted
+their minimum wage above the federal floor set at that time.
 
-{r}
-# MINIMUM WAGE GRAPH
+```{r echo=FALSE, message=FALSE, results='asis'}
+library(knitr)
+library(readxl)
+library(dplyr)
+library(ggplot2)
+
+# Load Data 
+welfare_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")
+
+# Retrieving data for state minimum wage for 2001 - 2019
+filtered_minwage <- welfare_data %>%
+  select(state_name, year, `State Minimum Wage`) %>%
+  filter(year >= 2001, year <= 2019)
+
+##### State Minimum Wage #####
+# Filter data for 2001 and 2019
+minwage_2001 <- welfare_data %>%
+  filter(year == 2001) %>%
+  select(state_name, year, `State Minimum Wage`)
+
+minwage_2019 <- welfare_data %>%
+  filter(year == 2019) %>%
+  select(state_name, year, `State Minimum Wage`)
+
+# Rename the `State Minimum Wage` columns to reflect the year
+minwage_2001 <- minwage_2001 %>%
+  rename(MinWage2001 = `State Minimum Wage`)
+
+minwage_2019 <- minwage_2019 %>%
+  rename(MinWage2019 = `State Minimum Wage`)
+
+# Merge the 2001 and 2019 data by state_name
+minwage_change <- left_join(minwage_2001, minwage_2019, by = "state_name")
+
+# Calculate the change in minimum wage from 2001 to 2019
+minwage_change <- minwage_change %>%
+  mutate(ChangeInMinWage = MinWage2019 - MinWage2001) %>%
+  select(state_name, MinWage2001, MinWage2019, ChangeInMinWage)
+
+kable(minwage_change, format = "markdown", caption = "Change in State Minimum Wage from 2001 to 2019")
+
+```
 
@@
 
-The trend lines across the graphs indicate a general decline in birth rates.
-The decline is more pronounced in states that have implemented delay policies, suggesting a possible correlation between policy enactment and a faster reduction in birth rates.
+The trend lines across the graphs indicate a general decline in birth
+rates. The decline is more pronounced in states that have implemented
+delay policies, suggesting a possible correlation between policy
+enactment and a faster reduction in birth rates.
 
@@
 
-Birth rates vary significantly between states. For example, Arkansas (AR) shows a sharp decrease in birth rates compared to other states. In contrast, Pennsylvania (PA) has a gentler slope.
+Birth rates vary significantly between states. For example, Arkansas
+(AR) shows a sharp decrease in birth rates compared to other states. In
+contrast, Pennsylvania (PA) has a gentler slope.
 
-The range of birth rates in 2001 spans from approximately 17 per 1000 (excluding Utah (UT)) people in certain states to below 15 per 1000 in others, with a convergence of rates around 12 to 14 per 1000.
+The range of birth rates in 2001 spans from approximately 17 per 1000
+(excluding Utah (UT)) people in certain states to below 15 per 1000 in
+others, with a convergence of rates around 12 to 14 per 1000.
 
-The states in this graph are ones which had a policy change from no delay to delay which allows for the analysis of the before-and-after effects of policy implementation.
+The states in this graph are ones which had a policy change from no
+delay to delay which allows for the analysis of the before-and-after
+effects of policy implementation.
 
-For example, Arizona's (AZ) implementation year of 2009 shows birth rate descends sharply afterward. In contrast, North Carolina (NC) in 2011 didn't show such an immediate shift.
+For example, Arizona's (AZ) implementation year of 2009 shows birth rate
+descends sharply afterward. In contrast, North Carolina (NC) in 2011
+didn't show such an immediate shift.
 
-The states without delay policies showcase a natural decline in birth rates over time, without the potential influence of policy changes.
-This trend is consistent across the states, with some variations in the steepness of the decline.
+The states without delay policies showcase a natural decline in birth
+rates over time, without the potential influence of policy changes. This
+trend is consistent across the states, with some variations in the
+steepness of the decline.
 
-For example, California (CA) maintains a relatively high birth rate over the years, starting near 15 per 1000 in 2001 and ending around 12 per 1000 in 2019.
-This could be influenced by its socio-economic profile and immigration patterns.
+For example, California (CA) maintains a relatively high birth rate over
+the years, starting near 15 per 1000 in 2001 and ending around 12 per
+1000 in 2019. This could be influenced by its socio-economic profile and
+immigration patterns.
 
-All states show a general decline in relative birth rates over the 19-year period, which is in line with the national trend of declining birth rates.
+All states show a general decline in relative birth rates over the
+19-year period, which is in line with the national trend of declining
+birth rates.
 
@@
 
-California (CA): The trend in California shows a relatively steady decline throughout the period, without any sharp changes.
+California (CA): The trend in California shows a relatively steady
+decline throughout the period, without any sharp changes.
 
-Florida (FL): Florida's trend is similar to California's, with a gradual decline.
+Florida (FL): Florida's trend is similar to California's, with a gradual
+decline.
 
-North Carolina (NC): North Carolina, with a policy change in 201, shows a relatively stable trend before the policy change and a sharper decline after.
+North Carolina (NC): North Carolina, with a policy change in 201, shows
+a relatively stable trend before the policy change and a sharper decline
+after.
 
-Texas (TX): Texas exhibited a noticeable decline after the policy implementation year of 2003.
+Texas (TX): Texas exhibited a noticeable decline after the policy
+implementation year of 2003.
 
-When contrasting the birth rates of Texas and California, two states with large populations, we see that Texas's birth rates declined more sharply after its policy change in 2003 compared to California.
-This might suggest an association between the policy implementation and the accelerated decline in birth rates in Texas.
+When contrasting the birth rates of Texas and California, two states
+with large populations, we see that Texas's birth rates declined more
+sharply after its policy change in 2003 compared to California. This
+might suggest an association between the policy implementation and the
+accelerated decline in birth rates in Texas.
 
@@
 
-Comparing Georgia and North Carolina, both of which implemented delay policies.
-Georgia's birth rates were rising before the policy change and then started declining after, while North Carolina's birth rates were relatively stable before 2011 and then declined.
+Comparing Georgia and North Carolina, both of which implemented delay
+policies. Georgia's birth rates were rising before the policy change and
+then started declining after, while North Carolina's birth rates were
+relatively stable before 2011 and then declined.
 
@@
 
-{r}
-#relative average # of births in states delay
+#relative average \# of births in states delay
 
-#relative average # of births in states w/o delay
+```{r echo=FALSE, message=FALSE}
+library(readxl)
+library(dplyr)
+library(readr)
+
+# Determine which states had a delay or no delay policy 
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(delay, stname, year)
+
+# Load the population data from the Excel file
+population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019)
+
+numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
+
+# Ensure both datasets have the 'stname' column for joining
+population_data <- population_data %>%
+  rename(stname = state_name)
+
+# Merge the number of births data with the population data
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname"))
+
+# Calculate the relative average number of births per 1000 people
+births_population_data <- births_population_data %>%
+  mutate(relative_births = (births / Population) * 1000)
+
+# Load the abortion policy data
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+  select(year, stname, delay)
+
+# Merge the relative births data with the abortion policy data
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
+
+# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
+states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
+states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
+
+# Filter the combined_data for states with no delay
+combined_data_no_delay <- combined_data %>%
+  filter(stname %in% states_no_delay)
+
+# Filter the combined_data for states with delay
+combined_data_with_delay <- combined_data %>%
+  filter(stname %in% states_with_delay)
+
+# Plot for states with policy delay
+ggplot(data = combined_data_with_delay) +
+  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
+  labs(title = "Relative Average Number of Births in States With Delay (2001-2019)",
+       x = "Year",
+       y = "Relative Average Number of Births per 1000 People",
+       color = "State") +
+  theme_minimal()
+
+```
+
+#relative average \# of births in states w/o delay
+
+```{r echo=FALSE, message=FALSE}
+library(readxl)
+library(dplyr)
+library(readr)
+
+# Determine which states had a delay or no delay policy 
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(delay, stname, year)
+
+# Load the population data from the Excel file
+population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019)
+
+numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
+
+# Ensure both datasets have the 'stname' column for joining
+population_data <- population_data %>%
+  rename(stname = state_name)
+
+# Merge the number of births data with the population data
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname"))
+
+# Calculate the relative average number of births per 1000 people
+births_population_data <- births_population_data %>%
+  mutate(relative_births = (births / Population) * 1000)
+
+# Load the abortion policy data
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+  select(year, stname, delay)
+
+# Merge the relative births data with the abortion policy data
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
+
+# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
+states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
+states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
+
+# Filter the combined_data for states with no delay
+combined_data_no_delay <- combined_data %>%
+  filter(stname %in% states_no_delay)
+
+# Filter the combined_data for states with delay
+combined_data_with_delay <- combined_data %>%
+  filter(stname %in% states_with_delay)
+
+# Plot for states with no policy delay
+ggplot(data = combined_data_no_delay) +
+  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
+  labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
+       x = "Year",
+       y = "Relative Average Number of Births per 1000 People",
+       color = "State") +
+  theme_minimal()
+
+```
+
+#relative birth rates in transition states
+
+```{r echo=FALSE, message=FALSE}
+library(readxl)
+library(dplyr)
+library(readr)
+
+# Determine which states had a delay or no delay policy 
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(delay, stname, year)
+
+# Load the population data from the Excel file
+population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019)
+
+numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
+
+# Ensure both datasets have the 'stname' column for joining
+population_data <- population_data %>%
+  rename(stname = state_name)
+
+# Merge the number of births data with the population data
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname"))
+
+# Calculate the relative average number of births per 1000 people
+births_population_data <- births_population_data %>%
+  mutate(relative_births = (births / Population) * 1000)
+
+# Load the abortion policy data
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+  select(year, stname, delay)
+
+# Merge the relative births data with the abortion policy data
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
+
+# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
+states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
+states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
+
+# Filter the combined_data for states with no delay
+combined_data_no_delay <- combined_data %>%
+  filter(stname %in% states_no_delay)
+
+# Filter the combined_data for states with delay
+combined_data_with_delay <- combined_data %>%
+  filter(stname %in% states_with_delay)
+
+# States and years of policy implementation
+policy_changes <- data.frame(
+  stname = c("AZ", "GA", "MN", "NC", "OK", "TX", "WV"),
+  year_implemented = c(2009, 2005, 2003, 2011, 2005, 2003, 2003)
+)
+
+# Filter combined_data
+transition_states_data <- combined_data %>%
+  filter(stname %in% policy_changes$stname)
+
+# Creating the graph
+ggplot(data = transition_states_data) +
+  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
+  geom_vline(data = policy_changes, aes(xintercept = year_implemented, color = stname), linetype="dashed") +
+  geom_text(data = policy_changes, aes(x = year_implemented, y = Inf, label = year_implemented), vjust = -0.5, color = "black") +
+  scale_x_continuous(breaks = seq(2001, 2019, by = 2)) +
+  labs(title = "Transition to Policy Delay and Relative Average Number of Births (2001-2019)",
+       x = "Year",
+       y = "Relative Average Number of Births per 1000 People",
+       color = "State") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
 
@@
 
+```{r echo=FALSE, message=FALSE}
+library(readxl)
+library(dplyr)
+library(readr)
+
+# Determine which states had a delay or no delay policy 
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(delay, stname, year)
+
+# Load the population data from the Excel file
+population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019)
+
+numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
+
+# Ensure both datasets have the 'stname' column for joining
+population_data <- population_data %>%
+  rename(stname = state_name)
+
+# Merge the number of births data with the population data
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname"))
+
+# Calculate the relative average number of births per 1000 people
+births_population_data <- births_population_data %>%
+  mutate(relative_births = (births / Population) * 1000)
+
+# Load the abortion policy data
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+  select(year, stname, delay)
+
+# Merge the relative births data with the abortion policy data
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
+
+# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
+states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
+states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
+
+# Filter the combined_data for states with no delay
+combined_data_no_delay <- combined_data %>%
+  filter(stname %in% states_no_delay)
+
+# Filter the combined_data for states with delay
+combined_data_with_delay <- combined_data %>%
+  filter(stname %in% states_with_delay)
+
+# Step 1: Select states for comparison
+states_comparison <- combined_data %>%
+  filter(stname %in% c("CA", "TX", "GA", "NC", "FL"))
+
+# Step 2: Calculate descriptive statistics and visualize trends
+# For California (no policy change) and Texas (policy change in 2003)
+ca_tx_data <- states_comparison %>%
+  filter(stname %in% c("CA", "TX"))
+
+# Calculate summary statistics for CA and TX before and after 2003
+ca_tx_summary <- ca_tx_data %>%
+  group_by(stname) %>%
+  summarise(
+    mean_before = mean(relative_births[year < 2003], na.rm = TRUE),
+    mean_after = mean(relative_births[year >= 2003], na.rm = TRUE)
+  )
+
+# For Georgia (policy change in 2005), North Carolina (policy change in 2011), and Florida (no policy change)
+ga_nc_fl_data <- states_comparison %>%
+  filter(stname %in% c("GA", "NC", "FL"))
+
+# Calculate summary statistics for GA and NC before and after policy changes, and compare with FL
+ga_nc_fl_summary <- ga_nc_fl_data %>%
+  group_by(stname) %>%
+  summarise(
+    mean_before = mean(relative_births[year < if_else(stname == "GA", 2005, if_else(stname == "NC", 2011, year))], na.rm = TRUE),
+    mean_after = mean(relative_births[year >= if_else(stname == "GA", 2005, if_else(stname == "NC", 2011, year))], na.rm = TRUE)
+  )
+
+# Step 3: Visualize trends
+ggplot(states_comparison, aes(x = year, y = relative_births, color = stname)) +
+  geom_line() +
+  geom_vline(xintercept = c(2003, 2005, 2011), linetype = "dashed", color = "black") +
+  labs(title = "Relative Birth Rates in Selected States (2001-2019)", y = "Relative Births per 1000 People", x = "Year") +
+  theme_minimal()
+
+```
+
 Second discussion: Political affiliation
 
-{r}
-# POLITICAL AFFILIATION MAP
+{r} \# POLITICAL AFFILIATION MAP
 
@@
 
-{r}
-#Education graph
+{r} #Education graph
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
+check. This shows...
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the
+prior. This shows...
 
-{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+{r} #\| eval: true #\| echo: false #\| message: false #\| warning: false
+#\| label: fig-ppcheckandposteriorvsprior #\| layout-ncol: 2 #\|
+fig-cap: "Examining how the model fits, and is affected by, the data"
+#\| fig-subcap: \["Posterior prediction check", "Comparing the posterior
+with the prior"\]
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
+pp_check(first_model) + theme_classic() + theme(legend.position =
+"bottom")
 
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
+posterior_vs_prior(first_model) + theme_minimal() +
+scale_color_brewer(palette = "Set1") + theme(legend.position =
+"bottom") + coord_flip()
 
@@
 
-{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
+{r} #\| echo: false #\| eval: true #\| message: false #\| warning: false
+#\| label: fig-stanareyouokay #\| fig-cap: "Checking the convergence of
+the MCMC algorithm" #\| fig-subcap: \["Trace plot", "Rhat plot"\] #\|
+layout-ncol: 2
 

### 2024-02-13 scripts/abortion_policy.R

@@ -1,24 +1,23 @@
 library(readxl)
 library(dplyr)
 library(readr)
-
+library(haven)
 
 # Determine which states had a delay or no delay policy 
-abortion_policy <- read_dta("annual_policy/policyvars01_19.dta") %>%
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
   filter(year >= 2001, year <= 2019) %>%
   select(delay, stname, year)
-view(abortion_policy)
 
 # Load the population data from the Excel file
-population_data <- read_excel("UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
+population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
   select(state_name, year, Population) %>%
   filter(year >= 2001, year <= 2019)
 
-numbirths_data <- read_csv("annual_policy/numbirths_2001_2019.csv") %>%
+numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
   rename(births = numbirth1544) %>%
   filter(year >= 2001, year <= 2019)
 
-# Ensure both datasets have the 'stname' column for joining
+# Ensure both data sets have the 'stname' column for joining
 population_data <- population_data %>%
   rename(stname = state_name)
 
@@ -31,7 +30,7 @@ births_population_data <- births_population_data %>%
   mutate(relative_births = (births / Population) * 1000)
 
 # Load the abortion policy data
-abortion_policy <- read_dta("annual_policy/policyvars01_19.dta") %>%
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
   select(year, stname, delay)
 
 # Merge the relative births data with the abortion policy data
@@ -50,7 +49,7 @@ combined_data_no_delay <- combined_data %>%
 combined_data_with_delay <- combined_data %>%
   filter(stname %in% states_with_delay)
 
-# Plot for states with no policy delay
+#### Plot for states with no policy delay ####
 ggplot(data = combined_data_no_delay) +
   geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
   labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
@@ -59,7 +58,7 @@ ggplot(data = combined_data_no_delay) +
        color = "State") +
[…]

### 2024-02-13 scripts/demographic_data.R

@@ -4,10 +4,10 @@ library(tidyr)
 
 
 # Importing the CSV files
-numbirths_2001_2019 <- read_csv("annual_policy/numbirths_2001_2019.csv")
-births_educ_race_age6 <- read_csv("decomp/births_educ_race_age6.csv")
-numbirths_educ_2044 <- read_csv("educ/numbirths_educ_2044.csv")
-state_births_04_19 <- read_csv("map/state_births_04_19.csv") 
+numbirths_2001_2019 <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv")
+births_educ_race_age6 <- read_csv("../data/analysis_data/decomp/births_educ_race_age6.csv")
+numbirths_educ_2044 <- read_csv("../data/analysis_data/educ/numbirths_educ_2044.csv")
+state_births_04_19 <- read_csv("../data/analysis_data/map/state_births_04_19.csv") 
 
 # Display the first few rows of each data-frame to verify the imports
 view(numbirths_2001_2019)
@@ -32,31 +32,7 @@ ggplot(average_births_by_state, aes(y = reorder(stname, AverageBirths), x = Aver
   labs(title = "Average Number of Births by State (2001-2019)",
        y = "State",
        x = "Average Number of Births") +
-  theme(axis.text.y = element_text(angle = 0)) # Ensure the y-axis labels are horizontal
-
-# Note: Customize the aesthetics as necessary to fit your dataset's specifics
+  theme(axis.text.y = element_text(angle = 0)) 
 
 
-
-
-
-
-### Number of Births for Education - REMOVE ### 
-filtered_numbirths_edu <- numbirths_educ_2044 %>%
-  filter(year >= 2001, year <= 2009)
-
-long_data <- filtered_numbirths_edu %>%
-  pivot_longer(cols = c("numbirth_lesshs", "numbirth_hsgrad", "numbirth_somecoll", "numbirth_colgrad", "numbirthalleduc", "numbirth_witheduc"),
-               names_to = "education_level",
-               values_to = "number_of_births")
-
-ggplot(long_data, aes(x = year, y = number_of_births, color = education_level, group = education_level)) +
-  geom_line() + 
-  geom_point() + 
-  theme_minimal() + 
-  labs(title = "Number of Births by Educational Attainment (2001-2009)",
-       x = "Year",
-       y = "Number of Births",
-       color = "Education Level") +
-  scale_x_continuous(breaks = 2001:2009) 
 

### 2024-02-13 scripts/fig1_replication.R

@@ -1,9 +1,9 @@
 # Step 1: Import Data
 library(readr)
-data_dir <- "/Users/mariamangru/Desktop/2023 - 2024 Academic Year/STA302/144981-V1"
+data_dir <- "../data/fig_1.csv"
 fig_1_data <- read_delim(file.path(data_dir, "fig_1.csv"), delim = ",", col_types = cols())
 
-# If any columns are read as character that should be numeric, convert them
+# If converting columns which are read as character to numeric 
 fig_1_data <- type.convert(fig_1_data, as.is = TRUE)
 
 # Step 2: Plotting Data
@@ -22,7 +22,7 @@ p <- ggplot(fig_1_data, aes(x = year, y = brate_all)) +
   annotate("text", x = 2007, y = min(fig_1_data$brate_all), label = "2007", vjust = -1)
 
 # Step 3: Export the Graph
-output_dir <- "/Users/mariamangru/Desktop/2023 - 2024 Academic Year/STA302/output"
+output_dir <- "../output"
 ggsave(file.path(output_dir, "fig_1.png"), plot = p, width = 10, height = 8, dpi = 300)
 
 

### 2024-02-13 scripts/minwage.R

@@ -3,7 +3,7 @@ library(dplyr)
 library(ggplot2)
 
 # Load Data 
-welfare_data <- read_excel("UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")
+welfare_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")
 View(welfare_data)
 
 colnames(welfare_data)

### 2024-02-13 scripts/regression_model.R

@@ -4,9 +4,9 @@ library(haven)
 library(tidyverse)
 
 # Reading the datasets
-abortion_policy <- read_dta("144981-V1/annual_policy/policyvars01_19.dta")
-numbirths <- read_csv("144981-V1/annual_policy/numbirths_2001_2019.csv")
-population_data <- read_excel("144981-V1/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")
+abortion_policy <- read_dta("..data/analysis_data/annual_policy/policyvars01_19.dta")
+numbirths <- read_csv("..data/analysis_data/annual_policy/numbirths_2001_2019.csv")
+population_data <- read_excel("..data/analysis_data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")
 
 # Filtering columns and filtering by year
 abortion_policy_filtered <- abortion_policy %>%

### 2024-02-13 child_expendure_graphs.qmd

---
title: "graph"
format: html
editor: visual
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Set Up

Setting up the library and getting the data.

```{r}
#### Workspace setup ####
library(tidyverse)
library(dplyr)
library(haven)
library(stata2r)

#### Download data ####

data <- read_dta("data/long_term/childcare_asec_09_1519.dta")
head(data)
```

You can add options to executable code like this

```{r}
group

ggplot(data, aes(x=year, y=obs)) +
    geom_col() 
```

```{r}
#### Download data ####

child_enforce_data <- read_dta("data/annual_policy/policyvars01_19.dta")
head(child_enforce_data)

# medicaid benefits
ggplot(child_enforce_data, aes(x=year, y=aca_mcaid_exp)) +
    geom_col()

# max well fair benefits
ggplot(child_enforce_data, aes(x=year, y=logben3)) +
    geom_point()

#parental consent for abortion
#ggplot(child_enforce_data, aes(x = year, y = parental)) +
#  geom_bar(stat = "identity")

#ggplot(child_enforce_data, aes(x = parental)) +
#  geom_bar(position = "dodge")

# Bar plot
ggplot(child_enforce_data, aes(x = factor(year), fill = factor(parental))) +
  geom_bar(position = "stack") +
  labs(x = "Year", y = "Count", fill = "Interest", title = "Parental Abortion Intrest Distribution Over the Years")
```

### 2024-02-14 child_expendure_graphs.qmd
imporvedd graphs asthetics and code readability

@@
 ---
-title: "graph"
+title: "child expenditure"
 format: html
@@
 ---
-
-## Quarto
-
-Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
 
@@
 
-data <- read_dta("data/long_term/childcare_asec_09_1519.dta")
-head(data)
-```
-
-You can add options to executable code like this
-
-```{r}
-group
-
-ggplot(data, aes(x=year, y=obs)) +
-    geom_col() 
+child_enforce_data <- read_dta("data/raw_data/annual_policy/policyvars01_19.dta")
+head(child_enforce_data)
 ```
@@
 ```{r}
-#### Download data ####
-
-child_enforce_data <- read_dta("data/annual_policy/policyvars01_19.dta")
-head(child_enforce_data)
 
 # medicaid benefits
-ggplot(child_enforce_data, aes(x=year, y=aca_mcaid_exp)) +
-    geom_col()
+ggplot(child_enforce_data, aes(x = year, y = aca_mcaid_exp)) +
+  geom_col(fill = "#1f77b4" ) + labs(x = "Year", y = "Number of people who have ACA Medicaid Expenditure", title = "People who have ACA Medicaid Expenditure Over Time (Years)") + theme_minimal() +  theme(plot.title = element_text(hjust = 0.5))
+```
 
+```{r}
 # max well fair benefits
 ggplot(child_enforce_data, aes(x=year, y=logben3)) +
-    geom_point()
+    geom_point()  + labs(x = "Year", y = "Max Welfare Benefits", title = "Max Welfare Benefits Over Time (Years)") + theme_minimal() +  theme(plot.title = element_text(hjust = 0.5))
 
-#parental consent for abortion
-#ggplot(child_enforce_data, aes(x = year, y = parental)) +
-#  geom_bar(stat = "identity")
+```
 
-#ggplot(child_enforce_data, aes(x = parental)) +
-#  geom_bar(position = "dodge")
-
-# Bar plot
+```{r}
+# Parental
 ggplot(child_enforce_data, aes(x = factor(year), fill = factor(parental))) +
   geom_bar(position = "stack") +
-  labs(x = "Year", y = "Count", fill = "Interest", title = "Parental Abortion Intrest Distribution Over the Years")
+  labs(x = "Year", y = "Parental Consent for Abortion", fill = "Interest", title = "Distribution of Interest in Abortion by Parent Over Time (Years)") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, vjust = 0.5))
 ```

### 2024-02-14 child_expendure_graphs.qmd
fixed medicaid graph y axisy

@@
 ggplot(child_enforce_data, aes(x = year, y = aca_mcaid_exp)) +
-  geom_col(fill = "#1f77b4" ) + labs(x = "Year", y = "Number of people who have ACA Medicaid Expenditure", title = "People who have ACA Medicaid Expenditure Over Time (Years)") + theme_minimal() +  theme(plot.title = element_text(hjust = 0.5))
+  geom_col(fill = "#1f77b4" ) + labs(x = "Year", y = "Number of people with ACA Medicaid Expenditure", title = "People who have ACA Medicaid Expenditure Over Time (Years)") + theme_minimal() +  theme(plot.title = element_text(hjust = 0.5))
 ```

### 2024-02-15 scripts/00-simulate_data.R
Simulate data, test simulated data, add literature, add sketch,

@@ -1,19 +1,51 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates some of our main data sets including birth rates per states, political affiliation per states and more.
+# Author: Rayan Awad Alim, Maria Mangu, Isha Junega
+# Date: 15 February 2024 
+# Contact: rayan.alim@mail.utoronto.ca, Maria.mangru@mail.utoronto.ca, isha.junega@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: N/A
 
 
-#### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
+#### Simulate Political Affiliation States data ####
+set.seed(23) 
 
-#### Simulate data ####
-# [...ADD CODE HERE...]
+#State Names
+states <- c("AK", "AL", "AR", "AS", "AZ", "CA", "CO", "CT", "DC", "DE", "FL", "GA", "GU", "HI",
+            "IA", "ID", "IL", "IN", "KS", "KY", "LA", "MA", "MD", "ME", "MI", "MN", "MO", "MP", 
+            "MS", "MT", "NC", "ND", "NE", "NH", "NJ", "NM", "NV", "NY", "OH", "OK", "OR", "PA",
+            "PR", "RI", "SC", "SD", "TN", "TX", "UM", "UT", "VA", "VI", "VT", "WA", "WI", "WV", "WY")
 
 
+# Simulate fraction of Senate that is Democrats ( between 0 and 1)
+fraction_of_senate_democrats <- runif(length(states))
 
+# Simulate birthrate per 1000 population between 5 and 20 (Typical range from literature)
+birthrate <- runif(length(states), min = 5, max = 20)
+
+#Craete dataframe of simulated dayta
+sim_datset <- data.frame(
+  State = states,
+  Fraction_of_senate_democrats = fraction_of_senate_democrats,
+  Birthrate = birthrate
+)
+
+head(sim_datset)
+
+## SOME TESTS FOR SIMULATED DATA ABOVE:
+# Test 1: Number of items matches the number of states
[…]

### 2024-02-15 scripts/abortion_policy.R
Multiple updates
Removed regression, changed abortion_policy, demographic_data and minwage scripts.

@@ -1,39 +1,46 @@
+#### Preamble ####
+# Purpose: Determine the effect of an abortion delay policy on birth rates
+# variables affect minimum wage 
+# Author: Maria Mangru
+# Date: 15th February, 2024
+# Contact: maria.mangru@utoronto.ca
+
+
+# Load libraries
 library(readxl)
 library(dplyr)
 library(readr)
 library(haven)
+library(ggplot2)
+library(lmtest)
+library(sandwich)
 
-# Determine which states had a delay or no delay policy 
-abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+# Define file paths 
+policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
+population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
+births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
+
+
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
   filter(year >= 2001, year <= 2019) %>%
-  select(delay, stname, year)
+  select(year, stname, delay)
 
-# Load the population data from the Excel file
-population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
-  select(state_name, year, Population) %>%
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
+  select(state_name = stname, year, Population) %>%
   filter(year >= 2001, year <= 2019)
 
-numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
   rename(births = numbirth1544) %>%
   filter(year >= 2001, year <= 2019)
 
-# Ensure both data sets have the 'stname' column for joining
-population_data <- population_data %>%
-  rename(stname = state_name)
-
-# Merge the number of births data with the population data
+# Merge the datasets
 births_population_data <- numbirths_data %>%
-  left_join(population_data, by = c("year", "stname"))
-
-# Calculate the relative average number of births per 1000 people
-births_population_data <- births_population_data %>%
+  left_join(population_data, by = c("year", "stname")) %>%
   mutate(relative_births = (births / Population) * 1000)
 
[…]

### 2024-02-15 scripts/fig1_replication.R

@@ -1,3 +1,10 @@
+#### Preamble ####
+# Purpose: Replication of figure 1
+# Author: Maria Mangru
+# Date: 15th February, 2024
+# Contact: maria.mangru@utoronto.ca
+
+
 # Step 1: Import Data
 library(readr)
 data_dir <- "../data/fig_1.csv"

### 2024-02-15 scripts/minwage.R

@@ -1,62 +1,94 @@
+#### Preamble ####
+# Purpose: Determines the impact of minimum wage on number of births 
+# Author: Maria Mangru
+# Date: 15th February, 2024
+# Contact: maria.mangru@utoronto.ca
+
+
+# Load libraries
 library(readxl)
+library(readr)
 library(dplyr)
 library(ggplot2)
+library(testthat)
 
-# Load Data 
-welfare_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")
-View(welfare_data)
-
-colnames(welfare_data)
-
-# Retrieving data for state minimum wage for 2001 - 2019
-filtered_minwage <- welfare_data %>%
+# Load the data sets
+welfare_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
   select(state_name, year, `State Minimum Wage`) %>%
-  filter(year >= 2001, year <= 2019)
+  filter(year >= 2001 & year <= 2019)
 
-view(filtered_minwage)
+numbirths_2001_2019 <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+  select(year, stname, numbirth1544) %>%
+  filter(year >= 2001 & year <= 2019) %>%
+  rename(state_name = stname)
+
+# Merge data sets on state_name and year
+merged_data <- merge(welfare_data, numbirths_2001_2019, by = c("state_name", "year"))
+
+# Calculate average minimum wage and average number of births per year 
+average_data <- merged_data %>%
+  group_by(year) %>%
+  summarise(Average_Minimum_Wage = mean(`State Minimum Wage`, na.rm = TRUE),
+            Average_Births = mean(numbirth1544, na.rm = TRUE))
 
 
-##### State Minimum Wage #####
-# Filter data for 2001 and 2019
-minwage_2001 <- welfare_data %>%
-  filter(year == 2001) %>%
-  select(state_name, year, `State Minimum Wage`)
+# Find maximum values for scaling 
+max_births <- max(average_data$Average_Births, na.rm = TRUE)
+max_wage <- max(average_data$Average_Minimum_Wage, na.rm = TRUE)
+scaling_factor <- max_wage / max_births
 
-minwage_2019 <- welfare_data %>%
-  filter(year == 2019) %>%
-  select(state_name, year, `State Minimum Wage`)
 
[…]

### 2024-02-15 scripts/regression_model.R

@@ -1,73 +1,0 @@
-library(readxl)
-library(readr)
-library(haven)
-library(tidyverse)
-
-# Reading the datasets
-abortion_policy <- read_dta("..data/analysis_data/annual_policy/policyvars01_19.dta")
-numbirths <- read_csv("..data/analysis_data/annual_policy/numbirths_2001_2019.csv")
-population_data <- read_excel("..data/analysis_data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")
-
-# Filtering columns and filtering by year
-abortion_policy_filtered <- abortion_policy %>%
-  select(year, stname, delay, aca_mcaid_exp, pcthsdrop, pcthsgrad, pctsomecol) %>%
-  filter(year >= 2001 & year <= 2019)
-
-population_data_filtered <- population_data %>%
-  select(state_name, year, 
-         Governor_is_Democrat = `Governor is Democrat (1=Yes)`, 
-         State_Minimum_Wage = `State Minimum Wage`) %>%
-  filter(year >= 2001 & year <= 2019) %>%
-  mutate(Governor_is_Democrat = na_if(Governor_is_Democrat, '.'))
-
-numbirths_filtered <- numbirths %>%
-  filter(year >= 2001 & year <= 2019)
-
-# Merging data sets and removing rows with NA values
-merged_data <- numbirths_filtered %>%
-  left_join(abortion_policy_filtered, by = c("year", "stname" = "stname")) %>%
-  left_join(population_data_filtered, by = c("stname" = "state_name", "year")) %>%
-  drop_na(Governor_is_Democrat, delay, aca_mcaid_exp, pcthsdrop, pcthsgrad, pctsomecol, numbirth1544, State_Minimum_Wage)
-
-# Running the regression model on the cleaned and filtered dataset
-model <- lm(numbirth1544 ~ delay + aca_mcaid_exp + pcthsdrop + pcthsgrad + pctsomecol + Governor_is_Democrat + State_Minimum_Wage, data = merged_data)
-
-# Summary of the model
-summary(model)
-
-
-###### Checking assumptions ######
-install.packages("car")
-install.packages("lmtest")
-install.packages("sandwich")
-
-library(tidyverse)
-library(car)
-library(lmtest)
-library(sandwich)
-
-
-# Check for linearity and homoscedasticity
-plot(model, which = 1:2)
-
-# Check for normality of residuals with a Q-Q plot
[…]

### 2024-02-15 paper/paper.qmd

@@
 
+# Workspace Set up
 library(tidyverse)
-
+library(haven)
+library(tidyr)
+library(readxl)
+library(dplyr)
+library(readr)
+library(knitr)
+library(ggplot2)
 ```
@@
 
-broader context to motivate The birth rate is one of the most important
-determinants of a country’s population growth, potentially impacting
-policy decisions about the health system, education, and economy (1).
+broader context to motivate The birth rate is one of the most important determinants of a country’s population growth, potentially impacting policy decisions about the health system, education, and economy (1).
 https://www.waldenu.edu/programs/health/resource/why-birth-rates-are-significant-in-health-studies-and-health-science
 
-some detail about what the paper is about This paper dives into the
-potential factors that may cause a birth rate to decrease. a clear gap
-that needs to be filled Not sure what was done We replicated the paper
-by Kearney, Levine, andPardue (2022) and focused on wages, access to
-healthcare (with religious importance) and childcare expenditure data to
-draw conclusions about whether these factors affect the decline or birth
-rates what was found
-
-why is it important the structure of the paper
+some detail about what the paper is about This paper dives into the potential factors that may cause a birth rate to decrease.
+a clear gap that needs to be filled Not sure what was done We replicated the paper by Kearney, Levine, andPardue (2022) and focused on wages, access to healthcare (with religious importance) and childcare expenditure data to draw conclusions about whether these factors affect the decline or birth rates what was found
 
@@
 
-Model Information Dependent variable: birth rate per 1000 population
-Independent variables:State min wage, Abortion delay, Child enforcement
-expenditure, Political affiliation
+Model InformationDependent variable: birth rate per 1000 populationIndependent variables:State min wage Abortion delay Child enforcement expenditure Political affiliation
 
@@
 ```
-We run the model in R [@citeR] using the rstanarm package of @rstanarm.
-We use the default priors from rstanarm.
+
+We run the model in R [@citeR] using the rstanarm package of @rstanarm. We use the default priors from rstanarm.
 
@@
 
-We expect a positive relationship between the size of the wings and time
-spent aloft. In particular...
+We expect a positive relationship between the size of the wings and time spent aloft. In particular...
 
-We can use maths by including latex between dollar signs, for instance
-$\theta$.
+We can use maths by including latex between dollar signs, for instance $\theta$.
 
@@
 
-Our results are summarized in @tbl-modelresults.
-
-```{r echo=FALSE, message=FALSE}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-
-```
-
 1- Medicaid coverage has increased which allows more low-income people
 to get an abortion.
+
+1- Medicaid coverage has increased which allows more low-income people to get an abortion.
+
+2- Maximum well-fare has gone down so it more difficult to raise a kid
+
+3- Parents seem to have more of an interest in having an abortion which aligns with the decrease in number of births
 
@@
 aligns with the decrease in number of births
-
-{r} #\| echo: false #\| eval: true #\| label: tbl-modelresults #\|
-tbl-cap: "Explanatory models of flight time based on wing width and wing
-length" #\| warning: false
-
-modelsummary::modelsummary( list( "First model" = first_model ),
-statistic = "mad", fmt = 2 )
 
@@
 ```{r echo=FALSE, message=FALSE}
-library(haven)
-library(tidyverse)
-library(tidyr)
 
@@
 ```{r echo=FALSE, message=FALSE, results='asis'}
-library(knitr)
-library(readxl)
-library(dplyr)
-library(ggplot2)
 
@@
 
-<!-- -->
-
 Comparing Georgia and North Carolina, both of which implemented delay
@@
 ```{r echo=FALSE, message=FALSE}
-library(readxl)
-library(dplyr)
-library(readr)
 
@@
 ```{r echo=FALSE, message=FALSE}
-library(readxl)
-library(dplyr)
-library(readr)
-
 # Determine which states had a delay or no delay policy 
@@
 
+Analysis Results
+
+### Analysis Results
+
+# State-Specific Birth Averages (2001-2019):
+
+The average number of births varies considerably from state to state, with California (CA) having the highest average number of births and Vermont (VT) having the lowest.
+
+Larger states, both in terms of geography and population like Texas (TX) and New York (NY), show significantly higher average numbers of births, which aligns with the expected outcomes based on population.
+
+Smaller states such as Vermont (VT) and North Dakota (ND) have lower average numbers of births, which may reflect not just smaller populations but also demographic trends such as aging populations.
+
+{r}
+#INTREST DISTRO OVER TIME GRAPHS
+
+# State Minimum Wage Changes (2001-2019):
+
+The District of Columbia (DC) exhibits the largest increase, with its minimum wage rising by $7.85, from $6.15 in 2001 to $14.00 in 2019.
+Arizona (AZ) and Colorado (CO) also show substantial increases of $6.85 each.
+
+Georgia (GA) and Wyoming (WY) stand out as the only states with no change in minimum wage, maintaining a rate of $5.15 across the years.
+It is worth noting that this rate is equal to the federal minimum wage between 1997 - 2007, suggesting that these states have not adjusted their minimum wage above the federal floor set at that time.
+
+{r}
+# MINIMUM WAGE GRAPH
+
+# State-Specific Births by Abortion Delay Policy from 2001 to 2019:
+
+The trend lines across the graphs indicate a general decline in birth rates.
+The decline is more pronounced in states that have implemented delay policies, suggesting a possible correlation between policy enactment and a faster reduction in birth rates.
+>>>>>>> 9e43bbea0ebd995129f3e0d14bc9e8ec625b3856
+
 #relative birth rates in selected states
@@
 
-{r} \# POLITICAL AFFILIATION MAP
+```{r}
+# POLITICAL AFFILIATION MAP
+```
+
+States with Delay
+
+Birth rates vary significantly between states. For example, Arkansas (AR) shows a sharp decrease in birth rates compared to other states. In contrast, Pennsylvania (PA) has a gentler slope.
+
+The range of birth rates in 2001 spans from approximately 17 per 1000 (excluding Utah (UT)) people in certain states to below 15 per 1000 in others, with a convergence of rates around 12 to 14 per 1000.
+
+The states in this graph are ones which had a policy change from no delay to delay which allows for the analysis of the before-and-after effects of policy implementation.
+
+For example, Arizona's (AZ) implementation year of 2009 shows birth rate descends sharply afterward. In contrast, North Carolina (NC) in 2011 didn't show such an immediate shift.
+
+The states without delay policies showcase a natural decline in birth rates over time, without the potential influence of policy changes.
+This trend is consistent across the states, with some variations in the steepness of the decline.
+
+For example, California (CA) maintains a relatively high birth rate over the years, starting near 15 per 1000 in 2001 and ending around 12 per 1000 in 2019.
+This could be influenced by its socio-economic profile and immigration patterns.
+
+All states show a general decline in relative birth rates over the 19-year period, which is in line with the national trend of declining birth rates.
+
+State-Specific Trends:
+
+California (CA): The trend in California shows a relatively steady decline throughout the period, without any sharp changes.
+
+Florida (FL): Florida's trend is similar to California's, with a gradual decline.
+
+North Carolina (NC): North Carolina, with a policy change in 201, shows a relatively stable trend before the policy change and a sharper decline after.
+
+Texas (TX): Texas exhibited a noticeable decline after the policy implementation year of 2003.
+
+When contrasting the birth rates of Texas and California, two states with large populations, we see that Texas's birth rates declined more sharply after its policy change in 2003 compared to California.
+This might suggest an association between the policy implementation and the accelerated decline in birth rates in Texas.
+
+Comparing Georgia and North Carolina, both of which implemented delay policies.
+Georgia's birth rates were rising before the policy change and then started declining after, while North Carolina's birth rates were relatively stable before 2011 and then declined.
+
+Discussion
+
+## First discussion point: Abortion 
+```{r}
+
+#relative average # of births in states delay
+
+#relative average # of births in states w/o delay
+
+#relative birth rates in selected states
+
+```
+
+#Second discussion: Political affiliation
+
+```{r}
+# POLITICAL AFFILIATION MAP
+```
 
@@
 
-{r} #Education graph
+```{r}
+#Education graph
+```
 
@@
 
-Posterior predictive check
+```{r}
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
-check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the
-prior. This shows...
-
-{r} #\| eval: true #\| echo: false #\| message: false #\| warning: false
-#\| label: fig-ppcheckandposteriorvsprior #\| layout-ncol: 2 #\|
-fig-cap: "Examining how the model fits, and is affected by, the data"
-#\| fig-subcap: \["Posterior prediction check", "Comparing the posterior
-with the prior"\]
-
-pp_check(first_model) + theme_classic() + theme(legend.position =
-"bottom")
-
-posterior_vs_prior(first_model) + theme_minimal() +
-scale_color_brewer(palette = "Set1") + theme(legend.position =
-"bottom") + coord_flip()
-
-Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-{r} #\| echo: false #\| eval: true #\| message: false #\| warning: false
-#\| label: fig-stanareyouokay #\| fig-cap: "Checking the convergence of
-the MCMC algorithm" #\| fig-subcap: \["Trace plot", "Rhat plot"\] #\|
-layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-
-\newpage
-
-References
+```

### 2024-02-15 paper/paper.qmd
Add political affiliation and education to paper

@@
 
+### Preparing Workspace
+# Install and load necessary packages
+if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
+if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
+if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
+if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
+if (!requireNamespace("knitr", quietly = TRUE)) install.packages("knitr")
+if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")
+if (!requireNamespace("maps", quietly = TRUE)) install.packages("maps")
+if (!requireNamespace("haven", quietly = TRUE)) install.packages("haven")
+if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
+if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
+
 # Workspace Set up
 library(tidyverse)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(knitr)
+library(readr)
+library(maps)
 library(haven)
-library(tidyr)
 library(readxl)
-library(dplyr)
-library(readr)
-library(knitr)
-library(ggplot2)
+library(here)
 ```
@@
 Model InformationDependent variable: birth rate per 1000 populationIndependent variables:State min wage Abortion delay Child enforcement expenditure Political affiliation
-
-Education
 
@@
 relatively stable before 2011 and then declined.
+
+# Education
+```{r echo=FALSE, message=FALSE}
+#Education
+
+# Load Education data
+educ_data <- read_csv("../data/analysis_data/educ/numbirths_educ_2044.csv")
+
+# Creating the plot
+ggplot(educ_data, aes(x = year)) +
+  geom_line(aes(y = numbirth_lesshs, color = "Less than HS"), linewidth = 1) +
+  geom_line(aes(y = numbirth_hsgrad, color = "HS Graduate"), linewidth = 1) +
+  geom_line(aes(y = numbirth_somecoll, color = "Some College"), linewidth = 1) +
+  geom_line(aes(y = numbirth_colgrad, color = "College Graduate"), linewidth = 1) +
+  geom_line(
+    aes(y = numbirthalleduc, color = "All Education Levels"),
+    size = 1,
+    linetype = "dashed"
+  ) +
+  scale_color_manual(
+    values = c(
+      "Less than HS" = "blue",
+      "HS Graduate" = "green",
+      "Some College" = "orange",
+      "College Graduate" = "red",
+      "All Education Levels" = "grey"
+      )
+  ) +
+  labs(
+    title = "Number of Births by Education Level Over Time",
+    x = "Year",
+    y = "Number of Births",
+    color = "Education Level"
+  ) +
+  theme_minimal() +
+  theme(legend.title = element_text(size = 12),
+        legend.text = element_text(size = 10))
+
+# Display the plot
+ggsave("births_by_education_level_over_time.png",
+       width = 10,
+       height = 6)
+
+```
+# State Political Affiliation and Birth Rates
+```{r echo=FALSE, message=FALSE}
+# https://stackoverflow.com/questions/29381652/plotting-us-heat-map-with-text
+
+#reading the UKNCPR Datat
+poldata <- read_csv(here("data", "analysis_data", "UKCPR_National_Welfare_Data_Update_020623.csv"))
+
+poldata <- poldata %>%
+  select(state_name,
+         year,
+         Population,
+         `Fraction of State Senate that is Democrat`)
+
+# Create a mapping of state abbreviations to full names
+state_abbrevs <- state.abb
+state_names <- state.name
+
+# Create a data frame for the mapping
+state_mapping <-
+  data.frame(abbrev = tolower(state_abbrevs),
+             name = tolower(state_names))
+
+#Get numbirths data
+numbirths_data <- read_csv(here("data", "analysis_data", "annual_policy", "numbirths_2001_2019.csv"))
+births_2019 <- numbirths_data %>% filter(year == 2019)
+
+# Convert state abbreviations to full names in your datasets
+numbirths_2019_fullname <- births_2019 %>%
+  mutate(stname = tolower(stname)) %>%
+  left_join(state_mapping, by = c("stname" = "abbrev")) %>%
+  select(-stname) %>%
+  rename(stname = name)
+
+poldata_2019 <- poldata %>% filter(year == 2019)
+
+poldata_2019_fullname <- poldata_2019 %>%
+  mutate(state_name = tolower(state_name)) %>%
+  left_join(state_mapping, by = c("state_name" = "abbrev")) %>%
+  select(-state_name) %>%
+  rename(state_name = name)
+
+# Merge poldata with numbirths_data using full state names
+data_merged <-
+  left_join(numbirths_2019_fullname,
+            poldata_2019_fullname,
+            by = c("stname" = "state_name"))
+
+states_map <- map_data("state")
+
+# Merge with states map data
+map_data_merged <-
+  left_join(states_map, data_merged, by = c("region" = "stname"))
+```
+
+```{r echo=FALSE, message=FALSE}
+# Plot
+ggplot(map_data_merged) +
+  geom_polygon(
+    aes(
+      x = long,
+      y = lat,
+      group = group,
+      fill = `Fraction of State Senate that is Democrat`
+    ),
+    color = "white"
+  ) +
+  geom_text(
+    data = map_data_merged %>% group_by(region) %>% summarise(
+      long = mean(long),
+      lat = mean(lat),
+      numbirths = first(births)
+    ),
+    aes(x = long, y = lat, label = numbirths),
+    size = 3,
+    check_overlap = TRUE
+  ) +
+  scale_fill_gradient(low = "red",
+                      high = "blue",
+                      name = "Democrat Fraction") +
+  coord_fixed(1.3) +
+  labs(title = "2019 Births and Democrat Fraction in State Senate by State") +
+  theme_void() +
+  theme(legend.position = "right")
+```
 

### 2024-02-15 paper/paper.qmd
update qmd

--- 2024-02-15 paper/paper.qmd
@@
 number-sections: true
-bibliography: references.bib
+#bibliography: references.bib
 editor: 
@@
 library(ggplot2)
+library(lmtest)
+library(sandwich)
 ```
@@
 
-broader context to motivate The birth rate is one of the most important determinants of a country’s population growth, potentially impacting policy decisions about the health system, education, and economy (1).
-https://www.waldenu.edu/programs/health/resource/why-birth-rates-are-significant-in-health-studies-and-health-science
+A country's birth rate is a key determinant of its population growth,
+which has a profound impact on policy pertaining to education, health,
+and the economy in general. With changing societal expectations and
+family dynamics, birth rates across developed countries have seemingly
+been declining in recent decades. The replacement birth rate is 2.1
+births per female, and significant deviations from this figure have
+consequences that affect countries in several important ways. A high
+birth rate means that society has to adapt to support a growing
+population, which places a stress on resources like food and housing.
+However, a low birth rate means that a society cannot sustain their
+population and elders may be viewed as “burdens”. The contributing
+factors that drive birth rates vary depending by country. This paper
+will focus on exploring findings related to declining birth rates in the
+United States.
 
-some detail about what the paper is about This paper dives into the potential factors that may cause a birth rate to decrease.
-a clear gap that needs to be filled Not sure what was done We replicated the paper by Kearney, Levine, andPardue (2022) and focused on wages, access to healthcare (with religious importance) and childcare expenditure data to draw conclusions about whether these factors affect the decline or birth rates what was found
+Melissa S. Kearney, Phillip B. Levine, and Luke Pardue, analyze
+potential reasons for a decline in birth rate in their paper, “The
+Puzzle of Falling US Birth Rates since the Great Recession.” In their
+analysis, examine birth rates by demographic groups defined by age,
+education, race and ethnicity, marital status, and birth parity inorder
+to find clues that serve as an explanation for the decline. They found
+factors pointing to the time and location do not serve as enough
+explanation to the decline in birthrates, although less quantifiable,
+women shifting their priorities to their aspirations for life instead of
+having more children is an important contributor. Their analysis does
+not explore solid evidence of US-specific policies or economic factors
+that can explain the depth of that decline(). We address this gap by
+replicating their paper and focusing on wages, access to healthcare
+(with religious importance) and childcare expenditure data to draw
+conclusions about whether these factors affect the decline or birth
+rates. We found that XX. This is important because. This paper is
+structured to include the data, results, and the discussion that include
+our analysis. Although the code used in the original paper is Strata, we
+use R to replicate some of its graphs, as well as to conduct analysis of
+our own.
 
 Data
+
+Melissa S. Kearney, Phillip B. Levine, and Luke Pardue's paper "The
+Puzzle of Falling US Birth Rates since the Great Recession" provided one
+of the datasets used in this analysis, notably the policyvars01_19
+dataset. This dataset presents a thorough summary of state-level
+policies that may have an impact on birth rates in the US from 2001 to
+2019 by assembling data on numerous policy variables from numerous
+sources.
 
@@
 
-Model InformationDependent variable: birth rate per 1000 populationIndependent variables:State min wage Abortion delay Child enforcement expenditure Political affiliation
+As part of the investigation, the researchers investigated the impact of
+the abortion policy delay on relative number of births per 1000 people
+using a Difference-in-Differences approach. This was done to compare the
+change in outcomes over time for the policy between a treatment group
+and a control group. In this paper, we looked at the effect of the
+abortion delay policy implementation in North Carolina in 2011. The
+control group of choice was Tennesse as this state did not implement an
+abortion delay policy during the years 2001 - 2019. Tennesse was also
+selected as the control group due to its geographical proximity to North
+Carolina and demographic similarities. This was done to satisfy the
+parallel trends assumption. The model employed is as follows:
 
-Education
+$[ \text{Relative Birthsit} = \beta_0 + \beta_1 (\text{Treatment}_i) + \beta_2(\text{PostPolicy}_t) + \beta_3(\text{Treatment} \times \text{PostPolicy}t) + \varepsilon{it} ]$
 
-Model set-up
+Where:
 
-Model defined as:
+-   Relative Births~it~ is the number of births per 1000 people in state
+    i at time t
 
-```{=tex}
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-```
+-   $Treatment_i$ is an indicator for weather state i is NC
 
-We run the model in R [@citeR] using the rstanarm package of @rstanarm. We use the default priors from rstanarm.
+-   $PostPolicy_t$ is an indicator for years after the policy change
 
-Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
+-   $Treatment_i$ x $PostPolicy_t$ is the interaction term
 
@@
 
-1- Medicaid coverage has increased which allows more low-income people to get an abortion.
+1- Medicaid coverage has increased which allows more low-income people
+to get an abortion.
 
@@
 
-3- Parents seem to have more of an interest in having an abortion which aligns with the decrease in number of births
+3- Parents seem to have more of an interest in having an abortion which
+aligns with the decrease in number of births
 
@@
 
-The average number of births varies considerably from state to state,
-with California (CA) having the highest average number of births and
-Vermont (VT) having the lowest.
-
-Larger states, both in terms of geography and population like Texas (TX)
-and New York (NY), show significantly higher average numbers of births,
-which aligns with the expected outcomes based on population.
-
-Smaller states such as Vermont (VT) and North Dakota (ND) have lower
-average numbers of births, which may reflect not just smaller
-populations but also demographic trends such as aging populations.
+Figure 1 presents a summary of the average number of scaled births by
+state over the years 2001 to 2109. Generally, it can be seen that
+certain states exhibit higher average number of births, while others,
+usually smaller states, exhibit lower number of births. The three states
+with the highest averages of births are California (CA), Texas (TX) and
+New York (NY).
 
@@
 
-# Importing the CSV files
+# Importing the CSV file
 numbirths_2001_2019 <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv")
-births_educ_race_age6 <- read_csv("../data/analysis_data/decomp/births_educ_race_age6.csv")
-numbirths_educ_2044 <- read_csv("../data/analysis_data/educ/numbirths_educ_2044.csv")
-state_births_04_19 <- read_csv("../data/analysis_data/map/state_births_04_19.csv")
 
-# Display the first few rows of each data-frame to verify the imports
-view(numbirths_2001_2019)
-view(births_educ_race_age6)
-view(numbirths_educ_2044)
-view(state_births_04_19)
-
-### Number of Births ### 
-colnames(numbirths_2001_2019)
-
+# Calculate average births scaled down by 1000 for visual representation
 average_births_by_state <- numbirths_2001_2019 %>%
@@
   group_by(stname) %>%
-  summarize(AverageBirths = mean(numbirth1544, na.rm = TRUE)) %>%
+  summarize(AverageBirths = mean(numbirth1544, na.rm = TRUE) / 1000) %>% # Scale down by a factor of 1000
   ungroup()
-view(average_births_by_state)
 
-ggplot(average_births_by_state, aes(y = reorder(stname, AverageBirths), x = AverageBirths)) +
+# Create the bar plot with scaled averages
+ggplot(average_births_by_state, aes(x = reorder(stname, -AverageBirths), y = AverageBirths)) +
   geom_bar(stat = "identity", fill = "steelblue") +
   theme_minimal() +
-  labs(title = "Average Number of Births by State (2001-2019)",
-       y = "State",
-       x = "Average Number of Births") +
-  theme(axis.text.y = element_text(angle = 0)) 
+  labs(title = "Scaled Average Number of Births by State (2001-2019)",
+       x = "State",
+       y = "Scaled Average Number of Births") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```
+
+# Child Expenditure
+
+Figure C2 The number of people with ACA Medicaid increased dramatically
+from 2000 to about 2014. Then, since 2014, it has slowly increased
+non-decreasingly till 2020.
+
+Figure C3 There has been a linear decrease of the maximum welfare
+benefits from 2000 to 2015. From 2015 to 2020 there has been a small
+rise in the maximum welfare benefits. Overall, there has been a loss in
+the maximum welfare benefits, despite the small rise.
+
+Figure C4 The blue portion of each column represents the number of
+parents who have consent for an abortion. The red portion represents the
+number of parents who don’t have consent for an abortion. There is a
+slight increase in consent for abortion from 2001 to 2019
 
@@
 
-The District of Columbia (DC) exhibits the largest increase, with its
-minimum wage rising by \$7.85, from \$6.15 in 2001 to \$14.00 in 2019.
-Arizona (AZ) and Colorado (CO) also show substantial increases of \$6.85
-each.
-
-Georgia (GA) and Wyoming (WY) stand out as the only states with no
-change in minimum wage, maintaining a rate of \$5.15 across the years.
-It is worth noting that this rate is equal to the federal minimum wage
-between 1997 - 2007, suggesting that these states have not adjusted
-their minimum wage above the federal floor set at that time.
+Figure 5 shows the trend of the average minimum wage across states
+depicted by the blue line and the average number of births represented
+by the red line, between the years 2001 to 2009. The trend for average
+minimum wage shows a noticeable increase over time. In contrast, the
+average number of births shows a fluctuating trend. Starting from 2001,
+the number of births gradually increase, with a peak occurring around
+2007. After this peak, there is a discernible decline in the average
+number of births.
 
@@
 
-# Load Data 
-welfare_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")
+# Load the data sets
+welfare_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
+  select(state_name, year, `State Minimum Wage`) %>%
+  filter(year >= 2001 & year <= 2019)
 
-# Retrieving data for state minimum wage for 2001 - 2019
-filtered_minwage <- welfare_data %>%
-  select(state_name, year, `State Minimum Wage`) %>%
-  filter(year >= 2001, year <= 2019)
+numbirths_2001_2019 <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+  select(year, stname, numbirth1544) %>%
+  filter(year >= 2001 & year <= 2019) %>%
+  rename(state_name = stname)
 
-##### State Minimum Wage #####
-# Filter data for 2001 and 2019
-minwage_2001 <- welfare_data %>%
-  filter(year == 2001) %>%
-  select(state_name, year, `State Minimum Wage`)
+# Merge data sets on state_name and year
+merged_data <- merge(welfare_data, numbirths_2001_2019, by = c("state_name", "year"))
 
-minwage_2019 <- welfare_data %>%
-  filter(year == 2019) %>%
-  select(state_name, year, `State Minimum Wage`)
+# Calculate average minimum wage and average number of births per year 
+average_data <- merged_data %>%
+  group_by(year) %>%
+  summarise(Average_Minimum_Wage = mean(`State Minimum Wage`, na.rm = TRUE),
+            Average_Births = mean(numbirth1544, na.rm = TRUE))
 
-# Rename the `State Minimum Wage` columns to reflect the year
-minwage_2001 <- minwage_2001 %>%
-  rename(MinWage2001 = `State Minimum Wage`)
+# Find maximum values for scaling 
+max_births <- max(average_data$Average_Births, na.rm = TRUE)
+max_wage <- max(average_data$Average_Minimum_Wage, na.rm = TRUE)
+scaling_factor <- max_wage / max_births
 
-minwage_2019 <- minwage_2019 %>%
-  rename(MinWage2019 = `State Minimum Wage`)
-
-# Merge the 2001 and 2019 data by state_name
-minwage_change <- left_join(minwage_2001, minwage_2019, by = "state_name")
-
-# Calculate the change in minimum wage from 2001 to 2019
-minwage_change <- minwage_change %>%
-  mutate(ChangeInMinWage = MinWage2019 - MinWage2001) %>%
-  select(state_name, MinWage2001, MinWage2019, ChangeInMinWage)
-
-kable(minwage_change, format = "markdown", caption = "Change in State Minimum Wage from 2001 to 2019")
+# Plot the graph with adjusted scales
+ggplot(average_data) +
+  geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", size = 1) +
+  geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", size = 1) +
+  scale_y_continuous(
+    "Minimum Wage",
+    sec.axis = sec_axis(~ . / scaling_factor, name = "Number of Births")
+  ) +
+  theme_minimal() +
+  labs(title = "Comparison of Average Minimum Wage and Birth Rates (2001-2009)",
+       x = "Year")
 
@@
 
-# State-Specific Births by Abortion Delay Policy from 2001 to 2019:
+# Impact of Abortion Delay Policy
 
-The trend lines across the graphs indicate a general decline in birth
-rates. The decline is more pronounced in states that have implemented
-delay policies, suggesting a possible correlation between policy
-enactment and a faster reduction in birth rates.
+Abortion delay policies refer to state-level regulations which stipulate
+a required waiting period between a woman’s initial consultation for an
+abortion and the procedure itself. This paper looks at states which have
+consistently implemented an abortion delay policy, those without this
+policy and those which have shifted from having no delay to have a delay
+policy between the years 2001 to 2019.
 
-States with Delay
+```{r echo=FALSE, message=FALSE, results='asis'}
+# Define file paths 
+policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
+population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
+births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
 
-Birth rates vary significantly between states. For example, Arkansas
-(AR) shows a sharp decrease in birth rates compared to other states. In
-contrast, Pennsylvania (PA) has a gentler slope.
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(year, stname, delay)
 
-The range of birth rates in 2001 spans from approximately 17 per 1000
-(excluding Utah (UT)) people in certain states to below 15 per 1000 in
-others, with a convergence of rates around 12 to 14 per 1000.
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
 
-The states in this graph are ones which had a policy change from no
-delay to delay which allows for the analysis of the before-and-after
-effects of policy implementation.
-
-For example, Arizona's (AZ) implementation year of 2009 shows birth rate
-descends sharply afterward. In contrast, North Carolina (NC) in 2011
-didn't show such an immediate shift.
-
-The states without delay policies showcase a natural decline in birth
-rates over time, without the potential influence of policy changes. This
-trend is consistent across the states, with some variations in the
-steepness of the decline.
-
-For example, California (CA) maintains a relatively high birth rate over
-the years, starting near 15 per 1000 in 2001 and ending around 12 per
-1000 in 2019. This could be influenced by its socio-economic profile and
-immigration patterns.
-
-All states show a general decline in relative birth rates over the
-19-year period, which is in line with the national trend of declining
-birth rates.
-
-State-Specific Trends:
-
-California (CA): The trend in California shows a relatively steady
-decline throughout the period, without any sharp changes.
-
-Florida (FL): Florida's trend is similar to California's, with a gradual
-decline.
-
-North Carolina (NC): North Carolina, with a policy change in 201, shows
-a relatively stable trend before the policy change and a sharper decline
-after.
-
-Texas (TX): Texas exhibited a noticeable decline after the policy
-implementation year of 2003.
-
-When contrasting the birth rates of Texas and California, two states
-with large populations, we see that Texas's birth rates declined more
-sharply after its policy change in 2003 compared to California. This
-might suggest an association between the policy implementation and the
-accelerated decline in birth rates in Texas.
-
-Comparing Georgia and North Carolina, both of which implemented delay
-policies. Georgia's birth rates were rising before the policy change and
-then started declining after, while North Carolina's birth rates were
-relatively stable before 2011 and then declined.
-
-Discussion
-
-## First discussion point: Abortion
-
-#relative average \# of births in states delay
-
-```{r echo=FALSE, message=FALSE}
-
-# Determine which states had a delay or no delay policy 
-abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
-  filter(year >= 2001, year <= 2019) %>%
-  select(delay, stname, year)
-
-# Load the population data from the Excel file
-population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
-  select(state_name, year, Population) %>%
-  filter(year >= 2001, year <= 2019)
-
-numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
   rename(births = numbirth1544) %>%
@@
 
-# Ensure both datasets have the 'stname' column for joining
-population_data <- population_data %>%
-  rename(stname = state_name)
-
-# Merge the number of births data with the population data
+# Merge the datasets
 births_population_data <- numbirths_data %>%
-  left_join(population_data, by = c("year", "stname"))
-
-# Calculate the relative average number of births per 1000 people
-births_population_data <- births_population_data %>%
+  left_join(population_data, by = c("year", "stname")) %>%
   mutate(relative_births = (births / Population) * 1000)
 
-# Load the abortion policy data
-abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
-  select(year, stname, delay)
-
-# Merge the relative births data with the abortion policy data
 combined_data <- abortion_policy %>%
@@
 
-# Filter the combined_data for states with no delay
+# Filter data for states with no delay and plot
 combined_data_no_delay <- combined_data %>%
@@
 
-# Filter the combined_data for states with delay
-combined_data_with_delay <- combined_data %>%
-  filter(stname %in% states_with_delay)
-
-# Plot for states with policy delay
-ggplot(data = combined_data_with_delay) +
-  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
-  labs(title = "Relative Average Number of Births in States With Delay (2001-2019)",
+ggplot(data = combined_data_no_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
+  geom_line() +
+  labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
        x = "Year",
@@
   theme_minimal()
-
 ```
 
-#relative average \# of births in states w/o delay
+Across the 19-year span, the data for states which have not implemented
+an abortion policy showcases a wide range of trends. Several states
+demonstrate a peak in birth rates at various points during the study
+period. While there is variation between the states, the overall
+trajectory indicated that the early to late 2000s were a period of
+higher birth rates, with a shift toward lower birth rates post 2010. It
+is important to note that this data in Figure 7 showcases the relative
+average number of births per 1000 people and fluctuates from 10 to
+approximately 16.
 
-```{r echo=FALSE, message=FALSE}
-# Determine which states had a delay or no delay policy 
-abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+```{r echo=FALSE, message=FALSE, results='asis'}
+# Define file paths 
+policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
+population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
+births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
+
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
   filter(year >= 2001, year <= 2019) %>%
-  select(delay, stname, year)
+  select(year, stname, delay)
 
-# Load the population data from the Excel file
-population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
   select(state_name, year, Population) %>%
-  filter(year >= 2001, year <= 2019)
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
 
-numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
   rename(births = numbirth1544) %>%
@@
 
-# Filter the combined_data for states with no delay
-combined_data_no_delay <- combined_data %>%
-  filter(stname %in% states_no_delay)
-
-# Filter the combined_data for states with delay
+# Filter data for states with delay and plot
 combined_data_with_delay <- combined_data %>%
@@
 
-# Plot for states with no policy delay
-ggplot(data = combined_data_no_delay) +
-  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
-  labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
+ggplot(data = combined_data_with_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
+  geom_line() +
+  labs(title = "Relative Average Number of Births in States With Delay (2001-2019)",
        x = "Year",
@@
   theme_minimal()
-
 ```
 
-#relative birth rates in transition states
+In contrast, figure \[\] looks at the trend of relative average number
+of births which have consistently implemented a delay policy. Similar to
+states which have not implemented a delay policy, birth rates for these
+states have generally decreased over the 19 year period. Over this
+period, the relative average number of births per 1000 people fluctuated
+significantly, ranging from around 10 to over 20.
 
-```{r echo=FALSE, message=FALSE}
+```{r echo=FALSE, message=FALSE, results='asis'}
 
-# Determine which states had a delay or no delay policy 
-abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
+# Define file paths 
+policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
+population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
+births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
+
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
   filter(year >= 2001, year <= 2019) %>%
-  select(delay, stname, year)
+  select(year, stname, delay)
 
-# Load the population data from the Excel file
-population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
   select(state_name, year, Population) %>%
-  filter(year >= 2001, year <= 2019)
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
 
-numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
   rename(births = numbirth1544) %>%
@@
 
-# Filter the combined_data for states with no delay
-combined_data_no_delay <- combined_data %>%
-  filter(stname %in% states_no_delay)
-
-# Filter the combined_data for states with delay
-combined_data_with_delay <- combined_data %>%
-  filter(stname %in% states_with_delay)
-
-# States and years of policy implementation
+# States which had a change in policy and plot
 policy_changes <- data.frame(
@@
 
-# Filter combined_data
 transition_states_data <- combined_data %>%
@@
 
-# Creating the graph
-ggplot(data = transition_states_data) +
-  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
+# Filter and plot for states which had a change in policy
+transition_states_data <- combined_data %>%
+  filter(stname %in% policy_changes$stname)
+
+ggplot(data = transition_states_data, aes(x = year, y = relative_births, group = stname, color = stname)) +
+  geom_line() +
   geom_vline(data = policy_changes, aes(xintercept = year_implemented, color = stname), linetype="dashed") +
@@
                        theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+Lastly, figure \[\] looks at states which transitioned from having no
+abortion delay to policy to implementing this policy at various years
+(indicated by the broken lines). A common trend among these states was
+an initial period of relatively higher birth rates in the early 2000s,
+with subsequent declines over the years. This trend also closely follows
+what observed for states which implemented an abortion delay policy and
+those which have not implemented such a policy.
+
+# Difference-in-Difference
+
+```{r echo=FALSE, message=FALSE, results='asis'}
+# Define file paths 
+policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
+population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
+births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
+
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(year, stname, delay)
+
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
+
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
+
+# Merge the datasets
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname")) %>%
+  mutate(relative_births = (births / Population) * 1000)
+
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
+
+# Difference-in-Difference Analysis for NC with a policy change in 2011 compared to a no-delay state
+did_data <- combined_data %>%
+  filter(stname %in% c("NC", "TN"), year >= 2001, year <= 2019) %>%
+  mutate(treatment = if_else(stname == "NC", 1, 0),
+         post_policy = if_else(year >= 2011, 1, 0),
+         treatment_x_post = treatment * post_policy)
+
+# Fit the DiD model
+did_model <- lm(relative_births ~ treatment + post_policy + treatment_x_post, data = did_data)
+
+# Summary of the DiD model and robust standard errors
+summary_did <- summary(did_model)
+robust_summary <- coeftest(did_model, vcov = vcovHC(did_model, type = "HC1"))
+
+# Extract coefficients and their standard errors
+coefficients <- coef(robust_summary)
+std_errors <- sqrt(diag(vcovHC(did_model, type = "HC1")))
+
+# Combine coefficients and standard errors into a data frame
+coef_df <- data.frame(Estimate = coefficients, `Std. Error` = std_errors)
+
+# Output the robust summary as a table
+kable(coef_df, format = "markdown")
 
@@
 
-Analysis Results
+These abortion delay policies created natural experiments which can be
+used to apply a Difference-in-Differences approach. The results from
+comparing North Carolina (treatment) with Tennessee (control) are as
+follow: - The intercept ($B_0$) estimate is 13.4345, suggesting that the
+baseline relative number of births per 1000 people is 13.4345 in the
+absence of treatment and before the policy change. - The treatment
+effect ($B_1$) is estimated at 0.4887, which is statistically
+significant (p = 0.01078). This indicates that, prior to the policy
+change, NC had 0.4887 more births per 1000 people than TN. - The
+post_policy effect ($B_2$) is -1.2372, also statistically significant (p
+= 1.26e-07), suggesting a general decrease in relative births per 1000
+people after 2011 in both states. - The interaction term ($B_3$), which
+represents the estimated effect of the policy change, is -0.7770, with a
+p-value of 0.00569, indicating that the policy delay in NC is associated
+with a reduction of 0.7770 births per 1000 people relative to TN after
+the policy implementation. - The model explains a significant portion of
+the variance in relative births per 1000 people ($R\^2$ = 0.8267),
+indicating a good fit.
 
-### Analysis Results
+# Education
 
-# State-Specific Birth Averages (2001-2019):
+#Politicial Affiliation
 
-The average number of births varies considerably from state to state, with California (CA) having the highest average number of births and Vermont (VT) having the lowest.
+Discussion \# Medicaid - remove heading after The Affordable Care
+(ACA)’s Medicaid is “intended to address systemic health inequalities
+for millions of Americans who lack health insurance” Medicaid is
+“expected to provide coverage to low-income individuals,” and also
+covers low-income pregnant women. Excluding the states where abortion is
+banned, Medicaid covers abortions, making it more accessible. According
+to Figure C1, More people continue to have medicaid as seen based on the
+trend from 2000 to 2020, making an affordable abortion more accessible.
+In Figure C3, we can see that there is an increase in parental consent
+for abortion. In other words, more parents are increasingly interested
+in getting an abortion every year. A potential motivator for parents to
+want an abortion is that as a low-income family, it is harder to take
+care of a larger family. Welfare is “federal government-sponsored
+assistance programs for individuals and families in need.” In Figure C2,
+it is evident that there is a decrease in maximum welfare benefits,
+making it less affordable for lower-income families to have a larger
+family size.The increasing demand for an abortion and the insurance
+covering it may contribute to a decline in birthrates.
 
-Larger states, both in terms of geography and population like Texas (TX) and New York (NY), show significantly higher average numbers of births, which aligns with the expected outcomes based on population.
+# Abortion policy - remove heading after
 
-Smaller states such as Vermont (VT) and North Dakota (ND) have lower average numbers of births, which may reflect not just smaller populations but also demographic trends such as aging populations.
+Through the difference-in-difference analysis, the results indicate a
+decrease of approximately 0.777 births per 1000 people in North Carolina
+post policy implementation relative to TN, which did not implement the
+abortion delay policy. This decrease is statistically significant,
+suggesting that implementing an abortion delay policy may have a real
+effect on birth rates.
 
-{r}
-#INTREST DISTRO OVER TIME GRAPHS
+The observed decrease in birth rates in North Carolina post 2011 could
+be partially explained by increased barriers to access as highlighted by
+de Londras et al. whose review supports the results derived in this
+paper. This review summarizes the effects of mandated waiting periods
+(MWPs) on outcomes connected to abortion. Their analysis suggests that
+MWPs may cause delays in obtaining abortion services, which may lower
+the number of births particularly in young, unmarried women. Noting the
+disproportionate impact on marginalized populations, such as teenagers,
+persons of color, and those with lower socioeconomic standing, they also
+draw attention to the broader human rights concerns of MWPs.
 
-# State Minimum Wage Changes (2001-2019):
-
-The District of Columbia (DC) exhibits the largest increase, with its minimum wage rising by $7.85, from $6.15 in 2001 to $14.00 in 2019.
-Arizona (AZ) and Colorado (CO) also show substantial increases of $6.85 each.
-
-Georgia (GA) and Wyoming (WY) stand out as the only states with no change in minimum wage, maintaining a rate of $5.15 across the years.
-It is worth noting that this rate is equal to the federal minimum wage between 1997 - 2007, suggesting that these states have not adjusted their minimum wage above the federal floor set at that time.
-
-{r}
-# MINIMUM WAGE GRAPH
-
-# State-Specific Births by Abortion Delay Policy from 2001 to 2019:
-
-The trend lines across the graphs indicate a general decline in birth rates.
-The decline is more pronounced in states that have implemented delay policies, suggesting a possible correlation between policy enactment and a faster reduction in birth rates.
->>>>>>> 9e43bbea0ebd995129f3e0d14bc9e8ec625b3856
-
-#relative birth rates in selected states
-
-```{r echo=FALSE, message=FALSE}
-
-# Determine which states had a delay or no delay policy 
-abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
-  filter(year >= 2001, year <= 2019) %>%
-  select(delay, stname, year)
-
-# Load the population data from the Excel file
-population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
-  select(state_name, year, Population) %>%
-  filter(year >= 2001, year <= 2019)
-
-numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
-  rename(births = numbirth1544) %>%
-  filter(year >= 2001, year <= 2019)
-
-# Ensure both datasets have the 'stname' column for joining
-population_data <- population_data %>%
-  rename(stname = state_name)
-
-# Merge the number of births data with the population data
-births_population_data <- numbirths_data %>%
-  left_join(population_data, by = c("year", "stname"))
-
-# Calculate the relative average number of births per 1000 people
-births_population_data <- births_population_data %>%
-  mutate(relative_births = (births / Population) * 1000)
-
-# Load the abortion policy data
-abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
-  select(year, stname, delay)
-
-# Merge the relative births data with the abortion policy data
-combined_data <- abortion_policy %>%
-  left_join(births_population_data, by = c("year", "stname"))
-
-# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
-states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
-states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
-
-# Filter the combined_data for states with no delay
-combined_data_no_delay <- combined_data %>%
-  filter(stname %in% states_no_delay)
-
-# Filter the combined_data for states with delay
-combined_data_with_delay <- combined_data %>%
-  filter(stname %in% states_with_delay)
-
-# Step 1: Select states for comparison
-states_comparison <- combined_data %>%
-  filter(stname %in% c("CA", "TX", "GA", "NC", "FL"))
-
-# Step 2: Calculate descriptive statistics and visualize trends
-# For California (no policy change) and Texas (policy change in 2003)
-ca_tx_data <- states_comparison %>%
-  filter(stname %in% c("CA", "TX"))
-
-# Calculate summary statistics for CA and TX before and after 2003
-ca_tx_summary <- ca_tx_data %>%
-  group_by(stname) %>%
-  summarise(
-    mean_before = mean(relative_births[year < 2003], na.rm = TRUE),
-    mean_after = mean(relative_births[year >= 2003], na.rm = TRUE)
-  )
-
-# For Georgia (policy change in 2005), North Carolina (policy change in 2011), and Florida (no policy change)
-ga_nc_fl_data <- states_comparison %>%
-  filter(stname %in% c("GA", "NC", "FL"))
-
-# Calculate summary statistics for GA and NC before and after policy changes, and compare with FL
-ga_nc_fl_summary <- ga_nc_fl_data %>%
-  group_by(stname) %>%
-  summarise(
-    mean_before = mean(relative_births[year < if_else(stname == "GA", 2005, if_else(stname == "NC", 2011, year))], na.rm = TRUE),
-    mean_after = mean(relative_births[year >= if_else(stname == "GA", 2005, if_else(stname == "NC", 2011, year))], na.rm = TRUE)
-  )
-
-# Step 3: Visualize trends
-ggplot(states_comparison, aes(x = year, y = relative_births, color = stname)) +
-  geom_line() +
-  geom_vline(xintercept = c(2003, 2005, 2011), linetype = "dashed", color = "black") +
-  labs(title = "Relative Birth Rates in Selected States (2001-2019)", y = "Relative Births per 1000 People", x = "Year") +
-  theme_minimal()
-
-```
+The results of the DiD found in this paper are supported by strong
+standard errors. However, we need to consider the limitations of the DiD
+approach as there are several other factors which may impact birth rates
+such as economic fluctuations and access to healthcare. Future
+iterations of this paper could expand on this research by conducting
+multi-state analysis for a variety of confounding factors.
 
@@
 
-States with Delay
-
-Birth rates vary significantly between states. For example, Arkansas (AR) shows a sharp decrease in birth rates compared to other states. In contrast, Pennsylvania (PA) has a gentler slope.
-
-The range of birth rates in 2001 spans from approximately 17 per 1000 (excluding Utah (UT)) people in certain states to below 15 per 1000 in others, with a convergence of rates around 12 to 14 per 1000.
-
-The states in this graph are ones which had a policy change from no delay to delay which allows for the analysis of the before-and-after effects of policy implementation.
-
-For example, Arizona's (AZ) implementation year of 2009 shows birth rate descends sharply afterward. In contrast, North Carolina (NC) in 2011 didn't show such an immediate shift.
-
-The states without delay policies showcase a natural decline in birth rates over time, without the potential influence of policy changes.
-This trend is consistent across the states, with some variations in the steepness of the decline.
-
-For example, California (CA) maintains a relatively high birth rate over the years, starting near 15 per 1000 in 2001 and ending around 12 per 1000 in 2019.
-This could be influenced by its socio-economic profile and immigration patterns.
-
-All states show a general decline in relative birth rates over the 19-year period, which is in line with the national trend of declining birth rates.
-
-State-Specific Trends:
-
-California (CA): The trend in California shows a relatively steady decline throughout the period, without any sharp changes.
-
-Florida (FL): Florida's trend is similar to California's, with a gradual decline.
-
-North Carolina (NC): North Carolina, with a policy change in 201, shows a relatively stable trend before the policy change and a sharper decline after.
-
-Texas (TX): Texas exhibited a noticeable decline after the policy implementation year of 2003.
-
-When contrasting the birth rates of Texas and California, two states with large populations, we see that Texas's birth rates declined more sharply after its policy change in 2003 compared to California.
-This might suggest an association between the policy implementation and the accelerated decline in birth rates in Texas.
-
-Comparing Georgia and North Carolina, both of which implemented delay policies.
-Georgia's birth rates were rising before the policy change and then started declining after, while North Carolina's birth rates were relatively stable before 2011 and then declined.
-
-Discussion
-
-## First discussion point: Abortion 
-```{r}
-
-#relative average # of births in states delay
-
-#relative average # of births in states w/o delay
-
-#relative birth rates in selected states
-
-```
-
-#Second discussion: Political affiliation
-
-```{r}
-# POLITICAL AFFILIATION MAP
-```
-
-Third discussion point: Minimum Wage
-
 Fourth discussion point: Education
@@
 
-Weaknesses and next steps
+Weaknesses and next steps A weakness of this paper is a lack of
+economical context. For example, we analyze our data independent of
+events like the Recession that happened in 2008. Without providing
+context to current events in the time period, our paper gives a weaker
+justification for the decline in birthrates. Next steps for improving
+our analysis include not only providing context for the events happening
+in the US, but focusing on the birth rates on a more micro level. Since
+each state within the US has its own laws, it will provide more insight
+to focus on the birth rates within each state. Additionally, to improve
+our analysis furthermore, we can compare the US to other countries that
+are similar from an economical standpoint and compare their birth rate
+to the US.
 

### 2024-02-15 scripts/abortion_policy.R

@@ -28,8 +28,10 @@ abortion_policy <- read_dta(policy_data_path) %>%
 
 # Load and prepare the population data
 population_data <- read_excel(population_data_path, sheet = "Data") %>%
-  select(state_name = stname, year, Population) %>%
-  filter(year >= 2001, year <= 2019)
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
+
 
 # Load and prepare the number of births data
 numbirths_data <- read_csv(births_data_path) %>%

### 2024-02-15 paper/paper.qmd
changes to paper

--- 2024-02-15 paper/paper.qmd
@@
   left_join(states_map, data_merged, by = c("region" = "stname"))
-```
 
-```{r echo=FALSE, message=FALSE}
 # Plot
@@
       lat = mean(lat),
-      numbirths = first(births)
+      numbirths = first(numbirth1544)
     ),

### 2024-02-15 paper/paper.qmd

--- 2024-02-15 paper/paper.qmd
@@
 
+### Preparing Workspace
+# Install and load necessary packages
+if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
+if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
+if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
+if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
+if (!requireNamespace("knitr", quietly = TRUE)) install.packages("knitr")
+if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")
+if (!requireNamespace("maps", quietly = TRUE)) install.packages("maps")
+if (!requireNamespace("haven", quietly = TRUE)) install.packages("haven")
+if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
+if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
+if (!requireNamespace("lmtest", quietly = TRUE)) install.packages("lmtest")
+if (!requireNamespace("sandwich", quietly = TRUE)) install.packages("sandwich")
+
 # Workspace Set up
 library(tidyverse)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(knitr)
+library(readr)
+library(maps)
 library(haven)
-library(tidyr)
 library(readxl)
-library(dplyr)
-library(readr)
-library(knitr)
-library(ggplot2)
+library(here)
 library(lmtest)
@@
 
+Model set-up
+=======
 Where:
@@
 births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
+```
+
+The range of birth rates in 2001 spans from approximately 17 per 1000
+(excluding Utah (UT)) people in certain states to below 15 per 1000 in
+others, with a convergence of rates around 12 to 14 per 1000.
+
+The states in this graph are ones which had a policy change from no
+delay to delay which allows for the analysis of the before-and-after
+effects of policy implementation.
+
+For example, Arizona's (AZ) implementation year of 2009 shows birth rate
+descends sharply afterward. In contrast, North Carolina (NC) in 2011
+didn't show such an immediate shift.
+
+The states without delay policies showcase a natural decline in birth
+rates over time, without the potential influence of policy changes. This
+trend is consistent across the states, with some variations in the
+steepness of the decline.
+
+For example, California (CA) maintains a relatively high birth rate over
+the years, starting near 15 per 1000 in 2001 and ending around 12 per
+1000 in 2019. This could be influenced by its socio-economic profile and
+immigration patterns.
+
+All states show a general decline in relative birth rates over the
+19-year period, which is in line with the national trend of declining
+birth rates.
+
+State-Specific Trends:
+
+California (CA): The trend in California shows a relatively steady
+decline throughout the period, without any sharp changes.
+
+Florida (FL): Florida's trend is similar to California's, with a gradual
+decline.
+
+North Carolina (NC): North Carolina, with a policy change in 201, shows
+a relatively stable trend before the policy change and a sharper decline
+after.
+
+Texas (TX): Texas exhibited a noticeable decline after the policy
+implementation year of 2003.
+
+When contrasting the birth rates of Texas and California, two states
+with large populations, we see that Texas's birth rates declined more
+sharply after its policy change in 2003 compared to California. This
+might suggest an association between the policy implementation and the
+accelerated decline in birth rates in Texas.
+
+Comparing Georgia and North Carolina, both of which implemented delay
+policies. Georgia's birth rates were rising before the policy change and
+then started declining after, while North Carolina's birth rates were
+relatively stable before 2011 and then declined.
+
+# Education
+
+```{r echo=FALSE, message=FALSE}
+#Education
+
+# Load Education data
+educ_data <- read_csv("../data/analysis_data/educ/numbirths_educ_2044.csv")
+
+# Creating the plot
+ggplot(educ_data, aes(x = year)) +
+  geom_line(aes(y = numbirth_lesshs, color = "Less than HS"), linewidth = 1) +
+  geom_line(aes(y = numbirth_hsgrad, color = "HS Graduate"), linewidth = 1) +
+  geom_line(aes(y = numbirth_somecoll, color = "Some College"), linewidth = 1) +
+  geom_line(aes(y = numbirth_colgrad, color = "College Graduate"), linewidth = 1) +
+  geom_line(
+    aes(y = numbirthalleduc, color = "All Education Levels"),
+    size = 1,
+    linetype = "dashed"
+  ) +
+  scale_color_manual(
+    values = c(
+      "Less than HS" = "blue",
+      "HS Graduate" = "green",
+      "Some College" = "orange",
+      "College Graduate" = "red",
+      "All Education Levels" = "grey"
+      )
+  ) +
+  labs(
+    title = "Number of Births by Education Level Over Time",
+    x = "Year",
+    y = "Number of Births",
+    color = "Education Level"
+  ) +
+  theme_minimal() +
+  theme(legend.title = element_text(size = 12),
+        legend.text = element_text(size = 10))
+
+# Display the plot
+ggsave("births_by_education_level_over_time.png",
+       width = 10,
+       height = 6)
+
+```
+
+# State Political Affiliation and Birth Rates
+```{r echo=FALSE, message=FALSE}
+# https://stackoverflow.com/questions/29381652/plotting-us-heat-map-with-text
+
+#reading the UKNCPR Datat
+poldata <- read_csv(here("data", "analysis_data", "UKCPR_National_Welfare_Data_Update_020623.csv"))
+
+poldata <- poldata %>%
+  select(state_name,
+         year,
+         Population,
+         `Fraction of State Senate that is Democrat`)
+
+# Create a mapping of state abbreviations to full names
+state_abbrevs <- state.abb
+state_names <- state.name
+
+# Create a data frame for the mapping
+state_mapping <-
+  data.frame(abbrev = tolower(state_abbrevs),
+             name = tolower(state_names))
+
+#Get numbirths data
+numbirths_data <- read_csv(here("data", "analysis_data", "annual_policy", "numbirths_2001_2019.csv"))
+births_2019 <- numbirths_data %>% filter(year == 2019)
+
+# Convert state abbreviations to full names in your datasets
+numbirths_2019_fullname <- births_2019 %>%
+  mutate(stname = tolower(stname)) %>%
+  left_join(state_mapping, by = c("stname" = "abbrev")) %>%
+  select(-stname) %>%
+  rename(stname = name)
+
+poldata_2019 <- poldata %>% filter(year == 2019)
+
+poldata_2019_fullname <- poldata_2019 %>%
+  mutate(state_name = tolower(state_name)) %>%
+  left_join(state_mapping, by = c("state_name" = "abbrev")) %>%
+  select(-state_name) %>%
+  rename(state_name = name)
+
+# Merge poldata with numbirths_data using full state names
+data_merged <-
+  left_join(numbirths_2019_fullname,
+            poldata_2019_fullname,
+            by = c("stname" = "state_name"))
+
+states_map <- map_data("state")
+
+# Merge with states map data
+map_data_merged <-
+  left_join(states_map, data_merged, by = c("region" = "stname"))
+
+# Plot
+ggplot(map_data_merged) +
+  geom_polygon(
+    aes(
+      x = long,
+      y = lat,
+      group = group,
+      fill = `Fraction of State Senate that is Democrat`
+    ),
+    color = "white"
+  ) +
+  geom_text(
+    data = map_data_merged %>% group_by(region) %>% summarise(
+      long = mean(long),
+      lat = mean(lat),
+      numbirths = first(numbirth1544)
+    ),
+    aes(x = long, y = lat, label = numbirths),
+    size = 3,
+    check_overlap = TRUE
+  ) +
+  scale_fill_gradient(low = "red",
+                      high = "blue",
+                      name = "Democrat Fraction") +
+  coord_fixed(1.3) +
+  labs(title = "2019 Births and Democrat Fraction in State Senate by State") +
+  theme_void() +
+  theme(legend.position = "right")
+```
+
+Discussion
+
+## First discussion point: Abortion
+
+#relative average \# of births in states delay
+
+```{r echo=FALSE, message=FALSE}
+# Determine which states had a delay or no delay policy 
+abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
 
@@
 
-Discussion \# Medicaid - remove heading after The Affordable Care
+### Discussion
+
+# Medicaid - remove heading after The Affordable Care
 (ACA)’s Medicaid is “intended to address systemic health inequalities
@@
 
-Second discussion: Political affiliation
+Second discussion: Political affiliation and birth rates
 
-```{r}
-# POLITICAL AFFILIATION MAP
-```
+From our analysis of political affiliations over the years, there are some relation between the two, especially in years following aggresive rhetorics or policies about reproductive rights.
+This indicates that periods following aggressive political rhetoric or significant policy changes regarding reproductive rights often coincide with noticeable changes in birth rates.
+This trend is observable across various states and is particularly pronounced in areas where political affiliation (# of democrats in state senate) change very differently from one side to the other.
+The rhetoric employed by political leaders and the nature of the policies they implement can have a substantial impact on individuals' choices regarding family planning.
+For instance, policies that limit access to reproductive healthcare, including contraception and abortion services, can lead to an increase in birth rates, especially in regions where alternative options are scarce.[@Nargund2009DecliningBirthRate]
+
+Conversely, political movements that advocate for and expand access to reproductive healthcare are often associated with a stabilization or decrease in birth rates.
+This can be attributed to increased autonomy over reproductive choices, allowing individuals to make informed decisions about if and when to have children.
 

### 2024-02-15 paper/paper.qmd
complete paper

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Melissa S. Kearney, Phillip B. Levine, and Luke Pardue in their paper, 
+'The Puzzle of Falling US Birth Rates since the Great Recession', analyze the 
+decreasing birth rates within the US. In this paper, we repicate use their data 
+and replicate their results of their study. While they focused on examining 
+birth rates by demographics, we focus on minimum wages, education, abortion 
+poilicy, and child support enforcement. The importance of our research is that 
+it highlights the underlying cause of why parents may not to have many children, 
+which leads to a decline in birthrates."
 format: pdf
@@
 library(sandwich)
+
+# manages graph placement
+knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
 ```
 
-Introduction
+# Introduction {#sec-intro}
 
@@
 analysis, examine birth rates by demographic groups defined by age,
-education, race and ethnicity, marital status, and birth parity inorder
+education, race and ethnicity, marital status, and birth parity in order
 to find clues that serve as an explanation for the decline. They found
@@
 not explore solid evidence of US-specific policies or economic factors
-that can explain the depth of that decline(). We address this gap by
-replicating their paper and focusing on wages, access to healthcare
-(with religious importance) and childcare expenditure data to draw
-conclusions about whether these factors affect the decline or birth
-rates. We found that XX. This is important because. This paper is
-structured to include the data, results, and the discussion that include
-our analysis. Although the code used in the original paper is Strata, we
-use R to replicate some of its graphs, as well as to conduct analysis of
-our own.
+that can explain the depth of that decline. We address this gap by
+replicating their paper and focusing on wages, abortion policies,
+education and childcare support enforcement data to draw conclusions
+about whether these factors affect the decline or birth rates.
 
-Data
+In this paper, we created a linear regression model to determine the
+impact of abortion policy on number of birthrates. We analyzed the
+affects of education on the number of birthrates by tracking the the
+birthrate for each education level separately. Additionally, to gage the
+impact of childcare support enforcement on the number of birthrates, we
+looked at how the maximum welfare benefits, including Medicaid, affect
+low-income individuals. Analyzing the wages, abortion policies,
+education, and childcare support important because they give deeper
+insights regarding why parents may not prefer to have many children and
+therefore may cause a decline in birthrates. The paper is structured to
+include the data, results, and the discussion that include our analysis.
+Although the code used in the original paper is Strata, we use R to
+replicate some of its graphs, as well as to conduct analysis of our own.
 
-Melissa S. Kearney, Phillip B. Levine, and Luke Pardue's paper "The
-Puzzle of Falling US Birth Rates since the Great Recession" provided one
-of the datasets used in this analysis, notably the policyvars01_19
-dataset. This dataset presents a thorough summary of state-level
-policies that may have an impact on birth rates in the US from 2001 to
-2019 by assembling data on numerous policy variables from numerous
-sources.
+# Data {#sec-data}
 
-Article:
+## Data Source and Methodology
 
-Model
+As part of our study, we looked closely at the variables influencing the decline in US birth rates.
+Our goal was to expand on and close the gaps left by Melissa S.
+Kearney, Phillip B.
+Levine, and Luke Pardue in their paper, "The Puzzle of Falling US Birth Rates since the Great Recession." The interaction between birth rates and childcare costs, healthcare availability, education, political affiliation, and incomes was the main focus of our research.
+We replicated many of the visuals and analyses from the original paper using R for data visualization and analysis, building on its conclusions through the use of a variety of datasets and approaches.
+The primary datasets for our study came from a variety of sources, such as the UK Center for Poverty Research National Welfare Database, the annual policy analysis data, and comprehensive birth records from 2001 to 2019.
+The Center for Poverty Research National Welfare Database [@ukcprWelfareData] at the University of Kentucky supplied more data on state-level laws, minimum wage levels, and political affiliations to supplement these original statistics.
+The UKCPR National Welfare Data which is a comprehensive dataset that has been compiled and annually updated by the University of Kentucky Center for Poverty Research (UKCPR).
+This database includes many variables such as population, wage levels, political landscape, dynamics of poverty, employment, and welfare in the United States at the state level.
+Covering a wide time span from 1980 to 2021, this dataset provides a longitudinal view of significant socio-economic indicators across the country.
+Building on earlier work by Kearney and Levine [@Kearney2022Puzzle], state-specific abortion delay law data was updated using the Guttmacher Institute's "Counseling and Waiting Periods for Abortion" database.
+The Kaiser Family Foundation's "Status of State Medicaid Expansion Decisions: Interactive Map" (KFF 2021) and a collection of birth statistics were the sources of information on Medicaid expansions.
+The methods used to gather the data made it possible to analyze state-level policies that potentially affect US birth rates in detail between 2001 and 2019.
+Through the integration of publicly accessible datasets and meticulously curated policy variables, our study offered a comprehensive understanding of the diverse factors influencing the fluctuations in US birth rates.
+
+## Variables and Summary Statisitcs
+With an emphasis on the years 2001 to 2019, we created and improved a collection
+of variables which are needed for studying the decrease in birth rates 
+throughout the United States. Our results are stronger because we were able to 
+record the short-term impacts of various different socioeconomic determinants 
+on birth rates during this time. The development of important variables, their 
+importance, storage formats, and the interactions between them are described in 
+detail below. Graphs, summary statistics, and comments are used to provide 
+context.
+
+The first main variable of interest is the delay variable which looks at the 
+Abortion Policy Delay. This binary variable indicated whether or not a stated 
+has implemented a delay in abortion policy. Our Difference-in-Differences (DiD) 
+analysis relied on this variable to evaluate the policy's effect on birth rates.
+Secondly relative_births was created to determine the relative birth rates of 
+each state by population in order to provide a common metric for comparing birth
+rates among states with different population sizes. This variable was created 
+by combining birth statistics with population estimates. State Minimum Wage 
+(min_wage) indicates the minimum wage for each state in a certain year. 
+The purpose of including this economic variable in the study of the correlation
+between birth rates and salary levels was to test the hypothesis that family
+planning choices could be influenced by greater salaries. Healthcare Access 
+(healthcare_access): An integer-based composite index based on the availability 
+of healthcare facilities and the expansion of Medicaid. This variable sought 
+to measure the ease of access to healthcare services, as this could have an 
+effect on birth rates by affecting the availability of prenatal and postnatal 
+care. Next, the variable childcare_exp quantified whether or not parents 
+received any support for their childcare expenses. Considering the substantial 
+expenses linked to childrearing, this variable was essential to comprehending 
+the financial influences on decisions about family size.
+
+High-level cleaning included resolving missing numbers, guaranteeing consistency
+across datasets, and filtering data to the years 2001-2019. Harmonizing 
+variables such as state_name made proper merging possible. To aid in analysis, 
+numerical variables were inspected for anomalies and, if required, normalized.
+
+Visual insights into the effects of abortion policy delays and economic factors
+on birth rates were provided via graphs showing the trend in relative birth 
+rates over time, by state, and by policy status. A summary of the distribution 
+and dispersion of the data was given via summary statistics. For example, the 
+distribution of min_wage brought attention to economic inequities, while the 
+average relative birth rate and its standard deviation showed variances between
+states and years.
+
+The associations between the important variables were shown using scatter plots
+and correlation analysis. For example, the relationship between min salary and 
+relative births was investigated in order to comprehend the impact of the 
+economy on family planning choices.
+
+## Measurements
+
+These datasets have several limitations even if they are quite useful for our 
+research. Firstly, reporting variations in birth rates can occur between states 
+and over time. These disparities could be the result of delayed data entry, 
+modifications to reporting requirements, or variations in data collection 
+techniques. These discrepancies could distort the birth rate estimates, 
+causing an overestimation or underestimating in particular states.
+
+Secondly, although the policy data is extensive, it might not fully account 
+for all state-level programs that have the potential to affect birth rates. 
+In addition, they do not give insight into variations within the policy. Birth
+trends may also be significantly influenced by policies that are not covered by 
+the dataset, such as regional healthcare access initiatives or educational 
+initiatives, but these factors are not taken into consideration in our research.
+To address these limitations, we used several strategies. To confirm the trends 
+found and guarantee consistency across sources, we cross-referenced our primary 
+data sources with supplementary datasets and research. Next, in order to 
+determine the potential impact of data mistakes, we performed sensitivity 
+analysis. We evaluated the robustness of our findings to differences in data 
+quality by correcting for any under- or over reporting in birth data.
+
+# Results: {#sec-results}
+
+## Model
 
@@
 
-Model set-up
-=======
+### Model set-up
+
 Where:
@@
 
-Results
+## Birth Rate Results
 
-1- Medicaid coverage has increased which allows more low-income people
-to get an abortion.
-
-1- Medicaid coverage has increased which allows more low-income people
-to get an abortion.
-
-2- Maximum well-fare has gone down so it more difficult to raise a kid
-
-3- Parents seem to have more of an interest in having an abortion which
-aligns with the decrease in number of births
-
-2- Maximum well-fare has gone down so it more difficult to raise a kid
-
-3- Parents seem to have more of an interest in having an abortion which
-aligns with the decrease in number of births
-
-Analysis Results
-
-### Analysis Results
-
-# State-Specific Birth Averages (2001-2019):
+### State-Specific Birth Averages (2001-2019):
 
@@
 ```{r echo=FALSE, message=FALSE}
+#| label: fig-1
 
@@
        y = "Scaled Average Number of Births") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
 ```
 
-# Child Expenditure
+## State Minimum Wage Changes (2001-2019):
 
-Figure C2 The number of people with ACA Medicaid increased dramatically
-from 2000 to about 2014. Then, since 2014, it has slowly increased
-non-decreasingly till 2020.
-
-Figure C3 There has been a linear decrease of the maximum welfare
-benefits from 2000 to 2015. From 2015 to 2020 there has been a small
-rise in the maximum welfare benefits. Overall, there has been a loss in
-the maximum welfare benefits, despite the small rise.
-
-Figure C4 The blue portion of each column represents the number of
-parents who have consent for an abortion. The red portion represents the
-number of parents who don’t have consent for an abortion. There is a
-slight increase in consent for abortion from 2001 to 2019
-
-# State Minimum Wage Changes (2001-2019):
-
-Figure 5 shows the trend of the average minimum wage across states
+Figure 2 shows the trend of the average minimum wage across states
 depicted by the blue line and the average number of births represented
@@
 ```{r echo=FALSE, message=FALSE, results='asis'}
+#| label: fig-2
 
@@
 ggplot(average_data) +
-  geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", size = 1) +
-  geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", size = 1) +
+  #geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", size = 1) +
+  #geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", size = 1) +
+  geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", linewidth = 1) +
+  geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", linewidth = 1) +
   scale_y_continuous(
@@
 
-# Impact of Abortion Delay Policy
+## Impact of Abortion Delay Policy
 
@@
 ```{r echo=FALSE, message=FALSE, results='asis'}
+#| label: fig-3
 # Define file paths 
@@
 births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
+
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(year, stname, delay)
+
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
+
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
+
+# Merge the datasets
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname")) %>%
+  mutate(relative_births = (births / Population) * 1000)
+
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
+
+# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
+states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
+states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
+
+# Filter data for states with no delay and plot
+combined_data_no_delay <- combined_data %>%
+  filter(stname %in% states_no_delay)
+
+ggplot(data = combined_data_no_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
+  geom_line() +
+  labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
+       x = "Year",
+       y = "Relative Average Number of Births per 1000 People",
+       color = "State") +
+  theme_minimal()
 ```
 
-The range of birth rates in 2001 spans from approximately 17 per 1000
-(excluding Utah (UT)) people in certain states to below 15 per 1000 in
-others, with a convergence of rates around 12 to 14 per 1000.
+Across the 19-year span, the data for states which have not implemented
+an abortion policy showcases a wide range of trends. Several states
+demonstrate a peak in birth rates at various points during the study
+period. While there is variation between the states, the overall
+trajectory indicated that the early to late 2000s were a period of
+higher birth rates, with a shift toward lower birth rates post 2010. It
+is important to note that this data in Figure 3 showcases the relative
+average number of births per 1000 people and fluctuates from 10 to
+approximately 16.
 
-The states in this graph are ones which had a policy change from no
-delay to delay which allows for the analysis of the before-and-after
-effects of policy implementation.
+```{r echo=FALSE, message=FALSE, results='asis'}
+#| label: fig-4
+# Define file paths 
+policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
+population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
+births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
 
-For example, Arizona's (AZ) implementation year of 2009 shows birth rate
-descends sharply afterward. In contrast, North Carolina (NC) in 2011
-didn't show such an immediate shift.
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(year, stname, delay)
 
-The states without delay policies showcase a natural decline in birth
-rates over time, without the potential influence of policy changes. This
-trend is consistent across the states, with some variations in the
-steepness of the decline.
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
 
-For example, California (CA) maintains a relatively high birth rate over
-the years, starting near 15 per 1000 in 2001 and ending around 12 per
-1000 in 2019. This could be influenced by its socio-economic profile and
-immigration patterns.
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
 
-All states show a general decline in relative birth rates over the
-19-year period, which is in line with the national trend of declining
-birth rates.
+# Merge the datasets
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname")) %>%
+  mutate(relative_births = (births / Population) * 1000)
 
-State-Specific Trends:
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
 
-California (CA): The trend in California shows a relatively steady
-decline throughout the period, without any sharp changes.
+# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
+states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
+states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
 
-Florida (FL): Florida's trend is similar to California's, with a gradual
-decline.
+# Filter data for states with delay and plot
+combined_data_with_delay <- combined_data %>%
+  filter(stname %in% states_with_delay)
 
-North Carolina (NC): North Carolina, with a policy change in 201, shows
-a relatively stable trend before the policy change and a sharper decline
-after.
+ggplot(data = combined_data_with_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
+  geom_line() +
+  labs(title = "Relative Average Number of Births in States With Delay (2001-2019)",
+       x = "Year",
+       y = "Relative Average Number of Births per 1000 People",
+       color = "State") +
+  theme_minimal()
+```
 
-Texas (TX): Texas exhibited a noticeable decline after the policy
-implementation year of 2003.
+In contrast, figure 4 looks at the trend of relative average number
+of births which have consistently implemented a delay policy. Similar to
+states which have not implemented a delay policy, birth rates for these
+states have generally decreased over the 19 year period. Over this
+period, the relative average number of births per 1000 people fluctuated
+significantly, ranging from around 10 to over 20.
 
-When contrasting the birth rates of Texas and California, two states
-with large populations, we see that Texas's birth rates declined more
-sharply after its policy change in 2003 compared to California. This
-might suggest an association between the policy implementation and the
-accelerated decline in birth rates in Texas.
+```{r echo=FALSE, message=FALSE, results='asis'}
+#| label: fig-5
 
-Comparing Georgia and North Carolina, both of which implemented delay
-policies. Georgia's birth rates were rising before the policy change and
-then started declining after, while North Carolina's birth rates were
-relatively stable before 2011 and then declined.
+# Define file paths 
+policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
+population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
+births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
 
-# Education
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(year, stname, delay)
+
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
+
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
+
+# Merge the datasets
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname")) %>%
+  mutate(relative_births = (births / Population) * 1000)
+
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
+
+# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
+states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
+states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
+
+# States which had a change in policy and plot
+policy_changes <- data.frame(
+  stname = c("AZ", "GA", "MN", "NC", "OK", "TX", "WV"),
+  year_implemented = c(2009, 2005, 2003, 2011, 2005, 2003, 2003)
+)
+
+transition_states_data <- combined_data %>%
+  filter(stname %in% policy_changes$stname)
+
+# Filter and plot for states which had a change in policy
+transition_states_data <- combined_data %>%
+  filter(stname %in% policy_changes$stname)
+
+ggplot(data = transition_states_data, aes(x = year, y = relative_births, group = stname, color = stname)) +
+  geom_line() +
+  geom_vline(data = policy_changes, aes(xintercept = year_implemented, color = stname), linetype="dashed") +
+  geom_text(data = policy_changes, aes(x = year_implemented, y = Inf, label = year_implemented), vjust = -0.5, color = "black") +
+  scale_x_continuous(breaks = seq(2001, 2019, by = 2)) +
+                       labs(title = "Transition to Policy Delay and Relative Average Number of Births (2001-2019)",
+                            x = "Year",
+                            y = "Relative Average Number of Births per 1000 People",
+                            color = "State") +
+                       theme_minimal() +
+                       theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+Lastly, figure 5 looks at states which transitioned from having no
+abortion delay to policy to implementing this policy at various years
+(indicated by the broken lines). A common trend among these states was
+an initial period of relatively higher birth rates in the early 2000s,
+with subsequent declines over the years. This trend also closely follows
+what observed for states which implemented an abortion delay policy and
+those which have not implemented such a policy.
+
+### Difference-in-Difference
+
+```{r echo=FALSE, message=FALSE, results='asis'}
+#| label: tbl-1
+# Define file paths 
+policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
+population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
+births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
+
+# Load and prepare the abortion policy data
+abortion_policy <- read_dta(policy_data_path) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  select(year, stname, delay)
+
+# Load and prepare the population data
+population_data <- read_excel(population_data_path, sheet = "Data") %>%
+  select(state_name, year, Population) %>%
+  filter(year >= 2001, year <= 2019) %>%
+  rename(stname = state_name)
+
+# Load and prepare the number of births data
+numbirths_data <- read_csv(births_data_path) %>%
+  rename(births = numbirth1544) %>%
+  filter(year >= 2001, year <= 2019)
+
+# Merge the datasets
+births_population_data <- numbirths_data %>%
+  left_join(population_data, by = c("year", "stname")) %>%
+  mutate(relative_births = (births / Population) * 1000)
+
+combined_data <- abortion_policy %>%
+  left_join(births_population_data, by = c("year", "stname"))
+
+# Difference-in-Difference Analysis for NC with a policy change in 2011 compared to a no-delay state
+did_data <- combined_data %>%
+  filter(stname %in% c("NC", "TN"), year >= 2001, year <= 2019) %>%
+  mutate(treatment = if_else(stname == "NC", 1, 0),
+         post_policy = if_else(year >= 2011, 1, 0),
+         treatment_x_post = treatment * post_policy)
+
+# Fit the DiD model
+did_model <- lm(relative_births ~ treatment + post_policy + treatment_x_post, data = did_data)
+
+# Summary of the DiD model and robust standard errors
+summary_did <- summary(did_model)
+robust_summary <- coeftest(did_model, vcov = vcovHC(did_model, type = "HC1"))
+
+# Extract coefficients and their standard errors
+coefficients <- coef(robust_summary)
+std_errors <- sqrt(diag(vcovHC(did_model, type = "HC1")))
+
+# Combine coefficients and standard errors into a data frame
+coef_df <- data.frame(Estimate = coefficients, `Std. Error` = std_errors)
+
+# Output the robust summary as a table
+kable(coef_df, format = "markdown")
+
+```
+
+These abortion delay policies created natural experiments which can be
+used to apply a Difference-in-Differences approach. The results from
+comparing North Carolina (treatment) with Tennessee (control) are as
+follow: - The intercept ($B_0$) estimate is 13.4345, suggesting that the
+baseline relative number of births per 1000 people is 13.4345 in the
+absence of treatment and before the policy change. - The treatment
+effect ($B_1$) is estimated at 0.4887, which is statistically
+significant (p = 0.01078). This indicates that, prior to the policy
+change, NC had 0.4887 more births per 1000 people than TN. - The
+post_policy effect ($B_2$) is -1.2372, also statistically significant (p
+= 1.26e-07), suggesting a general decrease in relative births per 1000
+people after 2011 in both states. - The interaction term ($B_3$), which
+represents the estimated effect of the policy change, is -0.7770, with a
+p-value of 0.00569, indicating that the policy delay in NC is associated
+with a reduction of 0.7770 births per 1000 people relative to TN after
+the policy implementation. - The model explains a significant portion of
+the variance in relative births per 1000 people ($R\^2$ = 0.8267),
+indicating a good fit.
+
+## Education
 
 ```{r echo=FALSE, message=FALSE}
+#| label: fig-6
+
 #Education
@@
     aes(y = numbirthalleduc, color = "All Education Levels"),
-    size = 1,
+    linewidth = 1,
     linetype = "dashed"
@@
 
-# State Political Affiliation and Birth Rates
+TODO: Results description for Education
+
+### State Political Affiliation and Birth Rates
+
 ```{r echo=FALSE, message=FALSE}
+#| label: fig-7
 # https://stackoverflow.com/questions/29381652/plotting-us-heat-map-with-text
@@
 
-Discussion
+TODO: Results description for Political Affiliation
 
-## First discussion point: Abortion
-
-#relative average \# of births in states delay
+## Child Support Enforcement
 
 ```{r echo=FALSE, message=FALSE}
-# Determine which states had a delay or no delay policy 
-abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
-
-# Load and prepare the abortion policy data
-abortion_policy <- read_dta(policy_data_path) %>%
-  filter(year >= 2001, year <= 2019) %>%
-  select(year, stname, delay)
-
-# Load and prepare the population data
-population_data <- read_excel(population_data_path, sheet = "Data") %>%
-  select(state_name, year, Population) %>%
-  filter(year >= 2001, year <= 2019) %>%
-  rename(stname = state_name)
-
-# Load and prepare the number of births data
-numbirths_data <- read_csv(births_data_path) %>%
-  rename(births = numbirth1544) %>%
-  filter(year >= 2001, year <= 2019)
-
-# Merge the datasets
-births_population_data <- numbirths_data %>%
-  left_join(population_data, by = c("year", "stname")) %>%
-  mutate(relative_births = (births / Population) * 1000)
-
-combined_data <- abortion_policy %>%
-  left_join(births_population_data, by = c("year", "stname"))
-
-# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
-states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
-states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
-
-# Filter data for states with no delay and plot
-combined_data_no_delay <- combined_data %>%
-  filter(stname %in% states_no_delay)
-
-ggplot(data = combined_data_no_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
-  geom_line() +
-  labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
-       x = "Year",
-       y = "Relative Average Number of Births per 1000 People",
-       color = "State") +
-  theme_minimal()
+child_enforce_data <- read_dta("../data/raw_data/annual_policy/policyvars01_19.dta")
 ```
 
-Across the 19-year span, the data for states which have not implemented
-an abortion policy showcases a wide range of trends. Several states
-demonstrate a peak in birth rates at various points during the study
-period. While there is variation between the states, the overall
-trajectory indicated that the early to late 2000s were a period of
-higher birth rates, with a shift toward lower birth rates post 2010. It
-is important to note that this data in Figure 7 showcases the relative
-average number of births per 1000 people and fluctuates from 10 to
-approximately 16.
+The number of people with ACA Medicaid increased dramatically
+from 2000 to about 2014. As figure 8 depicts, since 2014, the number of people 
+with Medicaid has slowly increased non-decreasingly till 2020. The ACA Medicaid 
+expansion has been adopted and implemented by states at different times, which 
+explains some of the dramatic increases within the data.
 
-```{r echo=FALSE, message=FALSE, results='asis'}
-# Define file paths 
-policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
-population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
-births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
-
-# Load and prepare the abortion policy data
-abortion_policy <- read_dta(policy_data_path) %>%
-  filter(year >= 2001, year <= 2019) %>%
-  select(year, stname, delay)
-
-# Load and prepare the population data
-population_data <- read_excel(population_data_path, sheet = "Data") %>%
-  select(state_name, year, Population) %>%
-  filter(year >= 2001, year <= 2019) %>%
-  rename(stname = state_name)
-
-# Load and prepare the number of births data
-numbirths_data <- read_csv(births_data_path) %>%
-  rename(births = numbirth1544) %>%
-  filter(year >= 2001, year <= 2019)
-
-# Merge the datasets
-births_population_data <- numbirths_data %>%
-  left_join(population_data, by = c("year", "stname")) %>%
-  mutate(relative_births = (births / Population) * 1000)
-
-combined_data <- abortion_policy %>%
-  left_join(births_population_data, by = c("year", "stname"))
-
-# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
-states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
-states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
-
-# Filter data for states with delay and plot
-combined_data_with_delay <- combined_data %>%
-  filter(stname %in% states_with_delay)
-
-ggplot(data = combined_data_with_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
-  geom_line() +
-  labs(title = "Relative Average Number of Births in States With Delay (2001-2019)",
-       x = "Year",
-       y = "Relative Average Number of Births per 1000 People",
-       color = "State") +
-  theme_minimal()
+```{r echo=FALSE, message=FALSE, fig.cap= "Expansion of Medicaid from 2000 to 2020"}
+#| label: fig-8
+#| 
+# medicaid benefits
+ggplot(child_enforce_data, aes(x = year, y = aca_mcaid_exp)) +
+  geom_col(fill = "#1f77b4" ) + labs(x = "Year", y = "Number of people with ACA Medicaid ", title = "People with ACA Medicaid Over Time (Years)") + theme_minimal() +  theme(plot.title = element_text(hjust = 0.5))
 ```
 
-In contrast, figure \[\] looks at the trend of relative average number
-of births which have consistently implemented a delay policy. Similar to
-states which have not implemented a delay policy, birth rates for these
-states have generally decreased over the 19 year period. Over this
-period, the relative average number of births per 1000 people fluctuated
-significantly, ranging from around 10 to over 20.
+In Figure 9, there has been a linear decrease of the maximum welfare benefits
+from 2000 to 2015. From 2015 to 2020 there has been a small rise in the maximum 
+welfare benefits. Overall, there has been a loss in the maximum welfare 
+benefits, despite the small rise.
 
-```{r echo=FALSE, message=FALSE, results='asis'}
-
-# Define file paths 
-policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
-population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
-births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
-
-# Load and prepare the abortion policy data
-abortion_policy <- read_dta(policy_data_path) %>%
-  filter(year >= 2001, year <= 2019) %>%
-  select(year, stname, delay)
-
-# Load and prepare the population data
-population_data <- read_excel(population_data_path, sheet = "Data") %>%
-  select(state_name, year, Population) %>%
-  filter(year >= 2001, year <= 2019) %>%
-  rename(stname = state_name)
-
-# Load and prepare the number of births data
-numbirths_data <- read_csv(births_data_path) %>%
-  rename(births = numbirth1544) %>%
-  filter(year >= 2001, year <= 2019)
-
-# Merge the datasets
-births_population_data <- numbirths_data %>%
-  left_join(population_data, by = c("year", "stname")) %>%
-  mutate(relative_births = (births / Population) * 1000)
-
-combined_data <- abortion_policy %>%
-  left_join(births_population_data, by = c("year", "stname"))
-
-# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
-states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
-states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")
-
-# States which had a change in policy and plot
-policy_changes <- data.frame(
-  stname = c("AZ", "GA", "MN", "NC", "OK", "TX", "WV"),
-  year_implemented = c(2009, 2005, 2003, 2011, 2005, 2003, 2003)
-)
-
-transition_states_data <- combined_data %>%
-  filter(stname %in% policy_changes$stname)
-
-# Filter and plot for states which had a change in policy
-transition_states_data <- combined_data %>%
-  filter(stname %in% policy_changes$stname)
-
-ggplot(data = transition_states_data, aes(x = year, y = relative_births, group = stname, color = stname)) +
-  geom_line() +
-  geom_vline(data = policy_changes, aes(xintercept = year_implemented, color = stname), linetype="dashed") +
-  geom_text(data = policy_changes, aes(x = year_implemented, y = Inf, label = year_implemented), vjust = -0.5, color = "black") +
-  scale_x_continuous(breaks = seq(2001, 2019, by = 2)) +
-                       labs(title = "Transition to Policy Delay and Relative Average Number of Births (2001-2019)",
-                            x = "Year",
-                            y = "Relative Average Number of Births per 1000 People",
-                            color = "State") +
-                       theme_minimal() +
-                       theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```{r echo=FALSE, message=FALSE}
+#| label: fig-9
+# max well fair benefits
+ggplot(child_enforce_data, aes(x=year, y=logben3)) +
+    geom_point()  + labs(x = "Year", y = "Max Welfare Benefits", title = "Max Welfare Benefits Over Time (Years)") + theme_minimal() +  theme(plot.title = element_text(hjust = 0.5))  
 ```
 
-Lastly, figure \[\] looks at states which transitioned from having no
-abortion delay to policy to implementing this policy at various years
-(indicated by the broken lines). A common trend among these states was
-an initial period of relatively higher birth rates in the early 2000s,
-with subsequent declines over the years. This trend also closely follows
-what observed for states which implemented an abortion delay policy and
-those which have not implemented such a policy.
+In Figure 10, the blue portion of each column represents the number of
+parents who have consent for an abortion. The red portion represents the
+number of parents who don’t have consent for an abortion. There is a
+slight increase in consent for abortion from 2001 to 2019.
 
-# Difference-in-Difference
-
-```{r echo=FALSE, message=FALSE, results='asis'}
-# Define file paths 
-policy_data_path <- "../data/analysis_data/annual_policy/policyvars01_19.dta"
-population_data_path <- "../data/UKCPR_National_Welfare_Data_Update_020623.xlsx"
-births_data_path <- "../data/analysis_data/annual_policy/numbirths_2001_2019.csv"
-
-# Load and prepare the abortion policy data
-abortion_policy <- read_dta(policy_data_path) %>%
-  filter(year >= 2001, year <= 2019) %>%
-  select(year, stname, delay)
-
-# Load and prepare the population data
-population_data <- read_excel(population_data_path, sheet = "Data") %>%
-  select(state_name, year, Population) %>%
-  filter(year >= 2001, year <= 2019) %>%
-  rename(stname = state_name)
-
-# Load and prepare the number of births data
-numbirths_data <- read_csv(births_data_path) %>%
-  rename(births = numbirth1544) %>%
-  filter(year >= 2001, year <= 2019)
-
-# Merge the datasets
-births_population_data <- numbirths_data %>%
-  left_join(population_data, by = c("year", "stname")) %>%
-  mutate(relative_births = (births / Population) * 1000)
-
-combined_data <- abortion_policy %>%
-  left_join(births_population_data, by = c("year", "stname"))
-
-# Difference-in-Difference Analysis for NC with a policy change in 2011 compared to a no-delay state
-did_data <- combined_data %>%
-  filter(stname %in% c("NC", "TN"), year >= 2001, year <= 2019) %>%
-  mutate(treatment = if_else(stname == "NC", 1, 0),
-         post_policy = if_else(year >= 2011, 1, 0),
-         treatment_x_post = treatment * post_policy)
-
-# Fit the DiD model
-did_model <- lm(relative_births ~ treatment + post_policy + treatment_x_post, data = did_data)
-
-# Summary of the DiD model and robust standard errors
-summary_did <- summary(did_model)
-robust_summary <- coeftest(did_model, vcov = vcovHC(did_model, type = "HC1"))
-
-# Extract coefficients and their standard errors
-coefficients <- coef(robust_summary)
-std_errors <- sqrt(diag(vcovHC(did_model, type = "HC1")))
-
-# Combine coefficients and standard errors into a data frame
-coef_df <- data.frame(Estimate = coefficients, `Std. Error` = std_errors)
-
-# Output the robust summary as a table
-kable(coef_df, format = "markdown")
-
+```{r echo=FALSE, message=FALSE}
+ #| label: fig-10
+# Parental
+ggplot(child_enforce_data, aes(x = factor(year), fill = factor(parental))) +
+  geom_bar(position = "stack") +
+  labs(x = "Year", y = "Parental Consent for Abortion", fill = "Interest", title = "Distribution of Interest in Abortion by Parent Over Time (Years)") + theme_minimal() +  # Apply minimal theme
+  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, vjust = 0.5))  
 ```
 
-These abortion delay policies created natural experiments which can be
-used to apply a Difference-in-Differences approach. The results from
-comparing North Carolina (treatment) with Tennessee (control) are as
-follow: - The intercept ($B_0$) estimate is 13.4345, suggesting that the
-baseline relative number of births per 1000 people is 13.4345 in the
-absence of treatment and before the policy change. - The treatment
-effect ($B_1$) is estimated at 0.4887, which is statistically
-significant (p = 0.01078). This indicates that, prior to the policy
-change, NC had 0.4887 more births per 1000 people than TN. - The
-post_policy effect ($B_2$) is -1.2372, also statistically significant (p
-= 1.26e-07), suggesting a general decrease in relative births per 1000
-people after 2011 in both states. - The interaction term ($B_3$), which
-represents the estimated effect of the policy change, is -0.7770, with a
-p-value of 0.00569, indicating that the policy delay in NC is associated
-with a reduction of 0.7770 births per 1000 people relative to TN after
-the policy implementation. - The model explains a significant portion of
-the variance in relative births per 1000 people ($R\^2$ = 0.8267),
-indicating a good fit.
+# Discussion:
 
-# Education
-
-#Politicial Affiliation
-
-### Discussion
-
-# Medicaid - remove heading after The Affordable Care
-(ACA)’s Medicaid is “intended to address systemic health inequalities
-for millions of Americans who lack health insurance” Medicaid is
-“expected to provide coverage to low-income individuals,” and also
-covers low-income pregnant women. Excluding the states where abortion is
-banned, Medicaid covers abortions, making it more accessible. According
-to Figure C1, More people continue to have medicaid as seen based on the
-trend from 2000 to 2020, making an affordable abortion more accessible.
-In Figure C3, we can see that there is an increase in parental consent
-for abortion. In other words, more parents are increasingly interested
-in getting an abortion every year. A potential motivator for parents to
-want an abortion is that as a low-income family, it is harder to take
-care of a larger family. Welfare is “federal government-sponsored
-assistance programs for individuals and families in need.” In Figure C2,
-it is evident that there is a decrease in maximum welfare benefits,
-making it less affordable for lower-income families to have a larger
-family size.The increasing demand for an abortion and the insurance
-covering it may contribute to a decline in birthrates.
-
-# Abortion policy - remove heading after
-
+## Abortion Policy
 Through the difference-in-difference analysis, the results indicate a
@@
 
-Second discussion: Political affiliation and birth rates
+## Second discussion: Political affiliation and birth rates
 
-From our analysis of political affiliations over the years, there are some relation between the two, especially in years following aggresive rhetorics or policies about reproductive rights.
-This indicates that periods following aggressive political rhetoric or significant policy changes regarding reproductive rights often coincide with noticeable changes in birth rates.
-This trend is observable across various states and is particularly pronounced in areas where political affiliation (# of democrats in state senate) change very differently from one side to the other.
-The rhetoric employed by political leaders and the nature of the policies they implement can have a substantial impact on individuals' choices regarding family planning.
-For instance, policies that limit access to reproductive healthcare, including contraception and abortion services, can lead to an increase in birth rates, especially in regions where alternative options are scarce.[@Nargund2009DecliningBirthRate]
+From our analysis of political affiliations over the years, there are
+some relation between the two, especially in years following aggresive
+rhetorics or policies about reproductive rights. This indicates that
+periods following aggressive political rhetoric or significant policy
+changes regarding reproductive rights often coincide with noticeable
+changes in birth rates. This trend is observable across various states
+and is particularly pronounced in areas where political affiliation (#
+of democrats in state senate) change very differently from one side to
+the other. The rhetoric employed by political leaders and the nature of
+the policies they implement can have a substantial impact on
+individuals' choices regarding family planning. For instance, policies
+that limit access to reproductive healthcare, including contraception
+and abortion services, can lead to an increase in birth rates,
+especially in regions where alternative options are
+scarce.[@Nargund2009DecliningBirthRate]
 
-Conversely, political movements that advocate for and expand access to reproductive healthcare are often associated with a stabilization or decrease in birth rates.
-This can be attributed to increased autonomy over reproductive choices, allowing individuals to make informed decisions about if and when to have children.
+Conversely, political movements that advocate for and expand access to
+reproductive healthcare are often associated with a stabilization or
+decrease in birth rates. This can be attributed to increased autonomy
+over reproductive choices, allowing individuals to make informed
+decisions about if and when to have children.
 
-Fourth discussion point: Education
+## Education
 
-```{r}
-#Education graph
-```
+## Child Support Enforcement
 
-Weaknesses and next steps A weakness of this paper is a lack of
-economical context. For example, we analyze our data independent of
-events like the Recession that happened in 2008. Without providing
-context to current events in the time period, our paper gives a weaker
-justification for the decline in birthrates. Next steps for improving
-our analysis include not only providing context for the events happening
-in the US, but focusing on the birth rates on a more micro level. Since
-each state within the US has its own laws, it will provide more insight
-to focus on the birth rates within each state. Additionally, to improve
-our analysis furthermore, we can compare the US to other countries that
-are similar from an economical standpoint and compare their birth rate
-to the US.
+(ACA)’s Medicaid is “intended to address systemic health inequalities
+for millions of Americans who lack health insurance” Medicaid is
+“expected to provide coverage to low-income individuals,” and also
+covers low-income pregnant women. Excluding the states where abortion is
+banned, Medicaid covers abortions, making it more accessible. According
+to Figure C1, More people continue to have medicaid as seen based on the
+trend from 2000 to 2020, making an affordable abortion more accessible.
+In Figure C3, we can see that there is an increase in parental consent
+for abortion. In other words, more parents are increasingly interested
+in getting an abortion every year. A potential motivator for parents to
+want an abortion is that as a low-income family, it is harder to take
+care of a larger family. Welfare is “federal government-sponsored
+assistance programs for individuals and families in need.” In Figure C2,
+it is evident that there is a decrease in maximum welfare benefits,
+making it less affordable for lower-income families to have a larger
+family size.The increasing demand for an abortion and the insurance
+covering it may contribute to a decline in birthrates.
 
-Weaknesses and next steps should also be included.
+## 4.2 Weakness and Next Steps
+
+A weakness of this paper is a lack of economical context. For example,
+we analyze our data independent of events like the Recession that
+happened in 2008. Without providing context to current events in the
+time period, our paper gives a weaker justification for the decline in
+birthrates. Next steps for improving our analysis include not only
+providing context for the events happening in the US, but focusing on
+the birth rates on a more micro level. Since each state within the US
+has its own laws, it will provide more insight to focus on the birth
+rates within each state. Additionally, to improve our analysis
+furthermore, we can compare the US to other countries that are similar
+from an economical standpoint and compare their birth rate to the US.
 
@@
 
-\appendix
-
-Appendix
-
-Additional data details
-
-Model details
-
-```{r}
-
-```
+# References

### 2024-02-15 paper/paper.qmd
adjusted title

@@
 ---
-title: "Red States = Red Flags?"
-subtitle: "An analysis of birth rate trends from 2001 to 2019 and an exploration of political affiliation of different states and Birth rates in the United States,"
+title: "The Decline of US Birth Rates: Analyzing Minimum Wages, Abortion Policies, Education, and Child Support Enforcement"
+subtitle:  "Reproduction of ‘The Puzzle of Falling US Birth Rates since the Great Recession’ (Kearney, Levine and, Pardue,2022)"
+
 author: 

### 2024-02-15 paper/paper.qmd
adjusted reference
graphs fixed

@@
   - Maria Mangru
-  - Isha Junega
+  - Isha Juneja
 thanks: "Code and data are available at: https://github.com/RayanAlim/Birth_Rate_Analysis"
@@
 include the data, results, and the discussion that include our analysis.
-Although the code used in the original paper is Strata, we use R to
-replicate some of its graphs, as well as to conduct analysis of our own.
+Although the code used in the original paper is Strata, we use R[@citeR]
+to replicate some of its graphs, as well as to conduct analysis of our
+own.
 
@@
 
-As part of our study, we looked closely at the variables influencing the decline in US birth rates.
-Our goal was to expand on and close the gaps left by Melissa S.
-Kearney, Phillip B.
-Levine, and Luke Pardue in their paper, "The Puzzle of Falling US Birth Rates since the Great Recession." The interaction between birth rates and childcare costs, healthcare availability, education, political affiliation, and incomes was the main focus of our research.
-We replicated many of the visuals and analyses from the original paper using R for data visualization and analysis, building on its conclusions through the use of a variety of datasets and approaches.
-The primary datasets for our study came from a variety of sources, such as the UK Center for Poverty Research National Welfare Database, the annual policy analysis data, and comprehensive birth records from 2001 to 2019.
-The Center for Poverty Research National Welfare Database [@ukcprWelfareData] at the University of Kentucky supplied more data on state-level laws, minimum wage levels, and political affiliations to supplement these original statistics.
-The UKCPR National Welfare Data which is a comprehensive dataset that has been compiled and annually updated by the University of Kentucky Center for Poverty Research (UKCPR).
-This database includes many variables such as population, wage levels, political landscape, dynamics of poverty, employment, and welfare in the United States at the state level.
-Covering a wide time span from 1980 to 2021, this dataset provides a longitudinal view of significant socio-economic indicators across the country.
-Building on earlier work by Kearney and Levine [@Kearney2022Puzzle], state-specific abortion delay law data was updated using the Guttmacher Institute's "Counseling and Waiting Periods for Abortion" database.
-The Kaiser Family Foundation's "Status of State Medicaid Expansion Decisions: Interactive Map" (KFF 2021) and a collection of birth statistics were the sources of information on Medicaid expansions.
-The methods used to gather the data made it possible to analyze state-level policies that potentially affect US birth rates in detail between 2001 and 2019.
-Through the integration of publicly accessible datasets and meticulously curated policy variables, our study offered a comprehensive understanding of the diverse factors influencing the fluctuations in US birth rates.
+As part of our study, we looked closely at the variables influencing the
+decline in US birth rates. Our goal was to expand on and close the gaps
+left by Melissa S. Kearney, Phillip B. Levine, and Luke Pardue in their
+paper, "The Puzzle of Falling US Birth Rates since the Great Recession."
+The interaction between birth rates and childcare costs, healthcare
+availability, education, political affiliation, and incomes was the main
+focus of our research. We replicated many of the visuals and analyses
+from the original paper using R for data visualization and analysis,
+building on its conclusions through the use of a variety of datasets and
+approaches. The primary datasets for our study came from a variety of
+sources, such as the UK Center for Poverty Research National Welfare
+Database, the annual policy analysis data, and comprehensive birth
+records from 2001 to 2019. The Center for Poverty Research National
+Welfare Database [@ukcprWelfareData] at the University of Kentucky
+supplied more data on state-level laws, minimum wage levels, and
+political affiliations to supplement these original statistics. The
+UKCPR National Welfare Data which is a comprehensive dataset that has
+been compiled and annually updated by the University of Kentucky Center
+for Poverty Research (UKCPR). This database includes many variables such
+as population, wage levels, political landscape, dynamics of poverty,
+employment, and welfare in the United States at the state level.
+Covering a wide time span from 1980 to 2021, this dataset provides a
+longitudinal view of significant socio-economic indicators across the
+country. Building on earlier work by Kearney and Levine
+[@Kearney2022Puzzle], state-specific abortion delay law data was updated
+using the Guttmacher Institute's "Counseling and Waiting Periods for
+Abortion" database. The Kaiser Family Foundation's "Status of State
+Medicaid Expansion Decisions: Interactive Map" (KFF 2021) and a
+collection of birth statistics were the sources of information on
+Medicaid expansions. The methods used to gather the data made it
+possible to analyze state-level policies that potentially affect US
+birth rates in detail between 2001 and 2019. Through the integration of
+publicly accessible datasets and meticulously curated policy variables,
+our study offered a comprehensive understanding of the diverse factors
+influencing the fluctuations in US birth rates.
 
 ## Variables and Summary Statisitcs
-With an emphasis on the years 2001 to 2019, we created and improved a collection
-of variables which are needed for studying the decrease in birth rates 
-throughout the United States. Our results are stronger because we were able to 
-record the short-term impacts of various different socioeconomic determinants 
-on birth rates during this time. The development of important variables, their 
-importance, storage formats, and the interactions between them are described in 
-detail below. Graphs, summary statistics, and comments are used to provide 
+
+With an emphasis on the years 2001 to 2019, we created and improved a
+collection of variables which are needed for studying the decrease in
+birth rates throughout the United States. Our results are stronger
+because we were able to record the short-term impacts of various
+different socioeconomic determinants on birth rates during this time.
+The development of important variables, their importance, storage
+formats, and the interactions between them are described in detail
+below. Graphs, summary statistics, and comments are used to provide
 context.
 
-The first main variable of interest is the delay variable which looks at the 
-Abortion Policy Delay. This binary variable indicated whether or not a stated 
-has implemented a delay in abortion policy. Our Difference-in-Differences (DiD) 
-analysis relied on this variable to evaluate the policy's effect on birth rates.
-Secondly relative_births was created to determine the relative birth rates of 
-each state by population in order to provide a common metric for comparing birth
-rates among states with different population sizes. This variable was created 
-by combining birth statistics with population estimates. State Minimum Wage 
-(min_wage) indicates the minimum wage for each state in a certain year. 
-The purpose of including this economic variable in the study of the correlation
-between birth rates and salary levels was to test the hypothesis that family
-planning choices could be influenced by greater salaries. Healthcare Access 
-(healthcare_access): An integer-based composite index based on the availability 
-of healthcare facilities and the expansion of Medicaid. This variable sought 
-to measure the ease of access to healthcare services, as this could have an 
-effect on birth rates by affecting the availability of prenatal and postnatal 
-care. Next, the variable childcare_exp quantified whether or not parents 
-received any support for their childcare expenses. Considering the substantial 
-expenses linked to childrearing, this variable was essential to comprehending 
-the financial influences on decisions about family size.
+The first main variable of interest is the delay variable which looks at
+the Abortion Policy Delay. This binary variable indicated whether or not
+a stated has implemented a delay in abortion policy. Our
+Difference-in-Differences (DiD) analysis relied on this variable to
+evaluate the policy's effect on birth rates. Secondly relative_births
+was created to determine the relative birth rates of each state by
+population in order to provide a common metric for comparing birth rates
+among states with different population sizes. This variable was created
+by combining birth statistics with population estimates. State Minimum
+Wage (min_wage) indicates the minimum wage for each state in a certain
+year. The purpose of including this economic variable in the study of
+the correlation between birth rates and salary levels was to test the
+hypothesis that family planning choices could be influenced by greater
+salaries. Healthcare Access (healthcare_access): An integer-based
+composite index based on the availability of healthcare facilities and
+the expansion of Medicaid. This variable sought to measure the ease of
+access to healthcare services, as this could have an effect on birth
+rates by affecting the availability of prenatal and postnatal care.
+Next, the variable childcare_exp quantified whether or not parents
+received any support for their childcare expenses. Considering the
+substantial expenses linked to childrearing, this variable was essential
+to comprehending the financial influences on decisions about family
+size.
 
-High-level cleaning included resolving missing numbers, guaranteeing consistency
-across datasets, and filtering data to the years 2001-2019. Harmonizing 
-variables such as state_name made proper merging possible. To aid in analysis, 
-numerical variables were inspected for anomalies and, if required, normalized.
+High-level cleaning included resolving missing numbers, guaranteeing
+consistency across datasets, and filtering data to the years 2001-2019.
+Harmonizing variables such as state_name made proper merging possible.
+To aid in analysis, numerical variables were inspected for anomalies
+and, if required, normalized.
 
-Visual insights into the effects of abortion policy delays and economic factors
-on birth rates were provided via graphs showing the trend in relative birth 
-rates over time, by state, and by policy status. A summary of the distribution 
-and dispersion of the data was given via summary statistics. For example, the 
-distribution of min_wage brought attention to economic inequities, while the 
-average relative birth rate and its standard deviation showed variances between
-states and years.
+Visual insights into the effects of abortion policy delays and economic
+factors on birth rates were provided via graphs showing the trend in
+relative birth rates over time, by state, and by policy status. A
+summary of the distribution and dispersion of the data was given via
+summary statistics. For example, the distribution of min_wage brought
+attention to economic inequities, while the average relative birth rate
+and its standard deviation showed variances between states and years.
 
-The associations between the important variables were shown using scatter plots
-and correlation analysis. For example, the relationship between min salary and 
-relative births was investigated in order to comprehend the impact of the 
-economy on family planning choices.
+The associations between the important variables were shown using
+scatter plots and correlation analysis. For example, the relationship
+between min salary and relative births was investigated in order to
+comprehend the impact of the economy on family planning choices.
 
@@
 
-These datasets have several limitations even if they are quite useful for our 
-research. Firstly, reporting variations in birth rates can occur between states 
-and over time. These disparities could be the result of delayed data entry, 
-modifications to reporting requirements, or variations in data collection 
-techniques. These discrepancies could distort the birth rate estimates, 
-causing an overestimation or underestimating in particular states.
+These datasets have several limitations even if they are quite useful
+for our research. Firstly, reporting variations in birth rates can occur
+between states and over time. These disparities could be the result of
+delayed data entry, modifications to reporting requirements, or
+variations in data collection techniques. These discrepancies could
+distort the birth rate estimates, causing an overestimation or
+underestimating in particular states.
 
-Secondly, although the policy data is extensive, it might not fully account 
-for all state-level programs that have the potential to affect birth rates. 
-In addition, they do not give insight into variations within the policy. Birth
-trends may also be significantly influenced by policies that are not covered by 
-the dataset, such as regional healthcare access initiatives or educational 
-initiatives, but these factors are not taken into consideration in our research.
-To address these limitations, we used several strategies. To confirm the trends 
-found and guarantee consistency across sources, we cross-referenced our primary 
-data sources with supplementary datasets and research. Next, in order to 
-determine the potential impact of data mistakes, we performed sensitivity 
-analysis. We evaluated the robustness of our findings to differences in data 
-quality by correcting for any under- or over reporting in birth data.
+Secondly, although the policy data is extensive, it might not fully
+account for all state-level programs that have the potential to affect
+birth rates. In addition, they do not give insight into variations
+within the policy. Birth trends may also be significantly influenced by
+policies that are not covered by the dataset, such as regional
+healthcare access initiatives or educational initiatives, but these
+factors are not taken into consideration in our research. To address
+these limitations, we used several strategies. To confirm the trends
+found and guarantee consistency across sources, we cross-referenced our
+primary data sources with supplementary datasets and research. Next, in
+order to determine the potential impact of data mistakes, we performed
+sensitivity analysis. We evaluated the robustness of our findings to
+differences in data quality by correcting for any under- or over
+reporting in birth data.
 
@@
        y = "Scaled Average Number of Births") +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1))
+  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=7))
 ```
@@
 # Plot the graph with adjusted scales
-ggplot(average_data) +
-  #geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", size = 1) +
-  #geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", size = 1) +
-  geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", linewidth = 1) +
-  geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", linewidth = 1) +
+ggplot(average_data, aes(x = year)) +
+  geom_line(aes(y = Average_Minimum_Wage), colour = "blue", linewidth = 1) +
+  geom_line(aes(y = Average_Births * scaling_factor), colour = "red", linewidth = 1) +
   scale_y_continuous(
@@
   theme_minimal() +
-  labs(title = "Comparison of Average Minimum Wage and Birth Rates (2001-2009)",
-       x = "Year")
+  labs(title = "Comparison of Average Minimum Wage and Birth Rates (2001-2019)",
+       x = "Year") +
+  theme(plot.title = element_text(size = 10))
 
@@
        x = "Year",
-       y = "Relative Average Number of Births per 1000 People",
-       color = "State") +
-  theme_minimal()
+       y = "Relative Average Number of Births per 1000 People") +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(size = 8, hjust = 0.5), 
+    axis.title.y = element_text(size = 8), 
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    legend.position = "left", 
+    legend.text = element_text(size = 6), 
+    legend.title = element_text(size = 6), 
+    legend.margin = margin(t = 5, r = 10, b = 0, l = 10) 
+  )
+
 ```
@@
        x = "Year",
-       y = "Relative Average Number of Births per 1000 People",
-       color = "State") +
-  theme_minimal()
+       y = "Relative Average Number of Births per 1000 People") +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(size = 10),  
+    axis.title.y = element_text(size = 10),  
+    axis.text.x = element_text(angle = 45, hjust = 1),  
+    legend.position = "right",  
+    legend.text = element_text(size = 8),  
+    legend.title = element_text(size = 10),  
+    legend.key.size = unit(0.1, 'cm'),  
+    legend.spacing.y = unit(0.1, 'cm')  
+  )
 ```
 
-In contrast, figure 4 looks at the trend of relative average number
-of births which have consistently implemented a delay policy. Similar to
+In contrast, figure 4 looks at the trend of relative average number of
+births which have consistently implemented a delay policy. Similar to
 states which have not implemented a delay policy, birth rates for these
@@
                        theme_minimal() +
-                       theme(axis.text.x = element_text(angle = 45, hjust = 1))
+                       theme(
+                         plot.title = element_text(size = 10),
+                         axis.title.y = element_text(size = 10),
+                         axis.text.x = element_text(angle = 45, hjust = 1)
+                         )
 ```
@@
 # Output the robust summary as a table
-kable(coef_df, format = "markdown")
+kable(coef_df, format = "markdown", caption = "Table 1")
 
@@
 
-The number of people with ACA Medicaid increased dramatically
-from 2000 to about 2014. As figure 8 depicts, since 2014, the number of people 
-with Medicaid has slowly increased non-decreasingly till 2020. The ACA Medicaid 
-expansion has been adopted and implemented by states at different times, which 
-explains some of the dramatic increases within the data.
+The number of people with ACA Medicaid increased dramatically from 2000
+to about 2014. As figure 8 depicts, since 2014, the number of people
+with Medicaid has slowly increased non-decreasingly till 2020. The ACA
+Medicaid expansion has been adopted and implemented by states at
+different times, which explains some of the dramatic increases within
+the data.
 
@@
 
-In Figure 9, there has been a linear decrease of the maximum welfare benefits
-from 2000 to 2015. From 2015 to 2020 there has been a small rise in the maximum 
-welfare benefits. Overall, there has been a loss in the maximum welfare 
-benefits, despite the small rise.
+In Figure 9, there has been a linear decrease of the maximum welfare
+benefits from 2000 to 2015. From 2015 to 2020 there has been a small
+rise in the maximum welfare benefits. Overall, there has been a loss in
+the maximum welfare benefits, despite the small rise.
 
@@
 ## Abortion Policy
+
 Through the difference-in-difference analysis, the results indicate a
@@
 
-A weakness of this paper is a lack of economical context. For example,
-we analyze our data independent of events like the Recession that
-happened in 2008. Without providing context to current events in the
-time period, our paper gives a weaker justification for the decline in
-birthrates. Next steps for improving our analysis include not only
-providing context for the events happening in the US, but focusing on
-the birth rates on a more micro level. Since each state within the US
+An improvement we can add to this paper is adding additional economical
+context along with our data such as investigating events like the
+Recession that took place in 2008, and the conflict regarding
+reproductive right policies. Without providing context to current events
+in the time period, our paper gives a weaker justification for the
+decline in birthrates. Next steps for improving our analysis include not
+only providing context for the events happening in the US, but focusing
+on the birth rates on a more micro level. Since each state within the US
 has its own laws, it will provide more insight to focus on the birth
@@
 # References
+Grundy E, Murphy M.
+Demography and public health.
+In: Detels R, Gulliford M, Karim QA, Tan CC, editors.
+Oxford textbook of global public health.
+6th ed.
+Oxford, England: Oxford University Press, 718–35.
+2015.

### 2024-02-15 paper/paper.qmd
submission

@@
 ```{r echo=FALSE, message=FALSE, results='asis'}
-#| label: tbl-1
+#| label: fig-11
 # Define file paths 
@@
 kable(coef_df, format = "markdown", caption = "Table 1")
+
+# Plotting for the DiD analysis
+ggplot(did_data, aes(x = year, y = relative_births, color = stname)) +
+  geom_line() +
+  geom_vline(xintercept = 2011, linetype="dashed", color = "red") + # Policy change year
+  labs(title = "Relative Births in NC (Treatment) and TN (Comparison) Over Time",
+       x = "Year", y = "Relative Births per 1000 People") +
+  theme_minimal() +
+                       theme(
+                         plot.title = element_text(size = 10)
+                        )
 
@@
 
-TODO: Results description for Education
+The figure 6 displays the trend in number of births segmented by education achievement from 1990 to 2019.
+The education levels all follow a similar trend with the exception of less than HS.
+For this education level, there is a decline in number of births post 2008 whereas the other education levels showed an increase in birth rates.
 
@@
 
-## Second discussion: Political affiliation and birth rates
+## Political affiliation and birth rates
 
@@
 ## Education
+We observe a convergence in birth rates across different educational categories based on the data for number of births by education level.
+This could be a reflection of larger societal shifts in family planning preferences and educational achievement.
+College graduates' stable fertility rate followed by a fall in births could mean that this group is delaying or reducing the size of their family, possibly for financial or career-related reasons.
+The overall decline in births over the past few years across all educational levels may reflect a general trend toward smaller families, or it may be the result of other outside factors not shown in the graph, such as shifting social norms, shifting economic conditions, or expanding access to reproductive healthcare.
 
@@
 banned, Medicaid covers abortions, making it more accessible. According
-to Figure C1, More people continue to have medicaid as seen based on the
+to figure 8, More people continue to have medicaid as seen based on the
 trend from 2000 to 2020, making an affordable abortion more accessible.
-In Figure C3, we can see that there is an increase in parental consent
+In figure 10, we can see that there is an increase in parental consent
 for abortion. In other words, more parents are increasingly interested
@@
 care of a larger family. Welfare is “federal government-sponsored
-assistance programs for individuals and families in need.” In Figure C2,
+assistance programs for individuals and families in need.” In figure 9,
 it is evident that there is a decrease in maximum welfare benefits,
@@
 # References
-Grundy E, Murphy M.
-Demography and public health.
-In: Detels R, Gulliford M, Karim QA, Tan CC, editors.
-Oxford textbook of global public health.
-6th ed.
-Oxford, England: Oxford University Press, 718–35.
-2015.
+@Dataset{ukcprWelfareData,
+  title = {UKCPR National Welfare Data, 1980-2021},
+  author = {{University of Kentucky Center for Poverty Research}},
+  year = {2023},
+  month = {Feb},
+  address = {Lexington, KY},
+  url = {http://ukcpr.org/resources/national-welfare-data%7D,
+  note = {Accessed: <Feb 15, 2024>},
+}
+
+@Manual{citeR,
+    title = {R: A Language and Environment for Statistical Computing},
+    author = {{R Core Team}},
+    organization = {R Foundation for Statistical Computing},
+    address = {Vienna, Austria},
+    year = {2023},
+    url = {https://www.R-project.org/},
+  }
+
+@InBook{Grundy2015Demography,
+  author    = {Emily Grundy and Michael Murphy},
+  editor    = {Roger Detels and Martin Gulliford and Quarraisha Abdool Karim and Chorh Chuan Tan},
+  title     = {Demography and public health},
+  booktitle = {Oxford Textbook of Global Public Health},
+  publisher = {Oxford University Press},
+  address   = {Oxford, England},
+  edition   = {6},
+  pages     = {718--735},
+  year      = {2015},
+}
+
+@Article{Kearney2022Puzzle,
+  author    = {Melissa S. Kearney and Phillip B. Levine and Luke Pardue},
+  title     = {The Puzzle of Falling US Birth Rates since the Great Recession},
+  journal   = {Journal of Economic Perspectives},
+  year      = {2022},
+  volume    = {36},
+  number    = {1},
+  pages     = {151--176},
+  doi       = {10.1257/jep.36.1.151},
+}
+
+@Article{thereferencecanbewhatever,
+    title = {Welcome to the {tidyverse}},
+    author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy D'Agostino McGowan and Romain François and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Lin Pedersen and Evan Miller and Stephan Milton Bache and Kirill Müller and Jeroen Ooms and David Robinson and Dana Paige Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
+    year = {2019},
+    journal = {Journal of Open Source Software},
+    volume = {4},
+    number = {43},
+    pages = {1686},
+    doi = {10.21105/joss.01686},
+  }
+
+  @Article{Nargund2009DecliningBirthRate,
+    title = {Declining birth rate in Developed Countries: A radical policy re-think is required},
+    author = {Nargund, G.},
+    year = {2009},
+    journal = {Facts Views Vis Obgyn},
+    volume = {1},
+    number = {3},
+    pages = {191-193},
+    PMID = {25489464},
+    PMCID = {PMC4255510},
+}
+
+@Manual{citeLahman,
+    title = {Lahman: Sean `Lahman' Baseball Database},
+    author = {Michael Friendly and Chris Dalzell and Martin Monkman and Dennis Murphy},
+    year = {2020},
+    note = {R package version 8.0-0},
+    url = {https://CRAN.R-project.org/package=Lahman},
+  }
+
+@article{gebru2021datasheets,
+  title={Datasheets for datasets},
+  author={Gebru, Timnit and Morgenstern, Jamie and Vecchione, Briana and Vaughan, Jennifer Wortman and Wallach, Hanna and Iii, Hal Daum{\'e} and Crawford, Kate},
+  journal={Communications of the ACM},
+  volume={64},
+  number={12},
+  pages={86--92},
+  year={2021},
+  publisher={ACM New York, NY, USA}
+}
+
+@Misc{rstanarm,
+    title = {rstanarm: {Bayesian} applied regression modeling via {Stan}},
+    author = {Ben Goodrich and Jonah Gabry and Imad Ali and Sam Brilleman},
+    note = {R package version 2.21.3},
+    year = {2022},
+    url = {https://mc-stan.org/rstanarm/},
+  }
+
+@Manual{palmerpenguins,
+    title = {palmerpenguins: Palmer Archipelago (Antarctica) penguin data},
+    author = {Allison Marie Horst and Alison Presmanes Hill and Kristen B Gorman},
+    year = {2020},
+    note = {R package version 0.1.0},
+    doi = {10.5281/zenodo.3960218},
+    url = {https://allisonhorst.github.io/palmerpenguins/},
+  }
+
+@Article{rohan,
+    title = {Welcome to the {tidyverse}},
+    author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy D'Agostino McGowan and Romain François and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Lin Pedersen and Evan Miller and Stephan Milton Bache and Kirill Müller and Jeroen Ooms and David Robinson and Dana Paige Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
+    year = {2019},
+    journal = {Journal of Open Source Software},
+    volume = {4},
+    number = {43},
+    pages = {1686},
+    doi = {10.21105/joss.01686},
+  }
+
+@Manual{,
+  title = {readxl: Read Excel Files},
+  author = {Hadley Wickham and Jennifer Bryan},
+  year = {2023},
+  note = {https://readxl.tidyverse.org, https://github.com/tidyverse/readxl},
+}
+
+@Article{,
+    title = {Welcome to the {tidyverse}},
+    author = {Hadley Wickham and Mara Averick and Jennifer Bryan and
+      Winston Chang and Lucy D'Agostino McGowan and Romain François and
+      Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester
+      and Max Kuhn and Thomas Lin Pedersen and Evan Miller and Stephan
+      Milton Bache and Kirill Müller and Jeroen Ooms and David Robinson
+      and Dana Paige Seidel and Vitalie Spinu and Kohske Takahashi and
+      Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
+    year = {2019},
+    journal = {Journal of Open Source Software},
+    volume = {4},
+    number = {43},
+    pages = {1686},
+    doi = {10.21105/joss.01686},
+  }
+
+@Article{,
+    title = {Diagnostic Checking in Regression Relationships},
+    author = {Achim Zeileis and Torsten Hothorn},
+    journal = {R News},
+    year = {2002},
+    volume = {2},
+    number = {3},
+    pages = {7--10},
+    url = {https://CRAN.R-project.org/doc/Rnews/},
+  }
+
+@Manual{sandwich,
+  title = {sandwich: Robust Covariance Matrix Estimators},
+  author = {Achim Zeileis},
+  year = {2022},
+  note = {R package version 3.0-1},
+  url = {https://CRAN.R-project.org/package=sandwich},
+}
+
+@Article{,
+    title = {Econometric Computing with {HC} and {HAC} Covariance
+      Matrix Estimators},
+    author = {Achim Zeileis},
+    journal = {Journal of Statistical Software},
+    year = {2004},
+    volume = {11},
+    number = {10},
+    pages = {1--17},
+    doi = {10.18637/jss.v011.i10},
+  }
+
+@Article{,
+    author = {Hadley Wickham},
+    title = {testthat: Get Started with Testing},
+    journal = {The R Journal},
+    year = {2011},
+    volume = {3},
+    pages = {5--10},
+    url =
+      {https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf},
+  }
+
+@Article{delondras2022impact,
+  title = {The impact of mandatory waiting periods on abortion-related outcomes: a synthesis of legal and health evidence},
+  author = {Fiona de Londras and Amanda Cleeve and Maria I. Rodriguez and Alana Farrell and Magdalena Furgalska and Antonella Lavelanet},
+  journal = {BMC Public Health},
+  volume = {22},
+  pages = {1232},
+  doi = {10.1186/s12889-022-13620-z},
+  url = {https://bmcpublichealth.biomedcentral.com/articles/10.1186/s12889-022-13620-z},
+}

### 2024-02-15 paper/paper.qmd
fixed

--- 2024-02-15 paper/paper.qmd
@@
 number-sections: true
-#bibliography: references.bib
+bibliography: references.bib
 editor: 
@@
 include the data, results, and the discussion that include our analysis.
-Although the code used in the original paper is Strata, we use R[@citeR]
+Although the code used in the original paper is Strata, we use R[@citeR], and other data exploration libraries[@knitr], [@rDplyr], [@rReadr], [@rGgplot2], [@rTidyverse] 
 to replicate some of its graphs, as well as to conduct analysis of our
@@
 # References
-Grundy E, Murphy M.
-Demography and public health.
-In: Detels R, Gulliford M, Karim QA, Tan CC, editors.
-Oxford textbook of global public health.
-6th ed.
-Oxford, England: Oxford University Press, 718–35.
-2015.

### 2024-02-15 paper/paper.qmd
final submission

@@
 knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
+#surpress error
+knitr::opts_chunk$set(warning = FALSE, message = FALSE)
 ```
@@
 
+Figure 11 presents a comparison of Tennessee (TN) and North Carolina (NC)'s relative birth rates per 1,000 people from the early 2000s to the present.
+The birth rates of both states show a downward trend over time, with NC having a higher birth rate than TN up until about the year shown by the vertical dashed line, which is the year the abortion delay policy was put into place.
+After this, NC's birth rates begin to decrease more sharply than Tennessee's, raising the possibility that the policy may have an effect.
+In the years following the policy change, the birth rate gap between the two states gets closer.
+Both states' general lowering trends are in line with the country's declining birth rate trend during this
+
 ## Education
@@
 
-```{r echo=FALSE, message=FALSE}
+```{r echo=FALSE, message=FALSE, }
 #| label: fig-7
@@
 ```
-
-TODO: Results description for Political Affiliation
 
